### 2024-12-02 paper/dump.qmd
Initial Commit
Initial Commit

---
title: "dump"
format: html
editor: visual
---

## Park and Recreation Facility Projects {#sec-parks}

The Park and Recreation Facility Projects dataset [@toronto_parks], published by the Parks, Forestry and Recreation Division and last updated on July 22, 2024, contains information about park and recreation facility projects in Toronto. This dataset includes details about the project name, location, ward number, and project status. Park and recreation facilities are essential public amenities that contribute to the quality of life and well-being of residents, providing spaces for recreation, leisure, and community engagement.

We choose this dataset to evaluate the how budget allocations are related to the number of park and recreation facility projects in each ward. The number of park projects can serve as a proxy for community development and public service enhancement, reflecting the need for investments in public amenities and infrastructure. By analyzing the relationship between park projects and budget allocations, we can gain insights into how public service enhancements influence the distribution of resources across the city. This will also help us understand how the city's budget decisions align with community development trends and public service needs in different wards, and determine if there are any disparities in resource allocation based on public service enhancements to certain wards.

```{r}
parks_data <- here::here("data/01-raw_data/Park and Recreation Facility Projects - 4326.csv")
parks <- read_csv(parks_data)

# create a graph of the number of parks by ward_number
parks |>
  count(ward_number) |>
  ggplot(aes(x = ward_number, y = n)) +
  geom_col() +
  labs(x = "Ward number", y = "Number of parks") +
  theme_minimal()
```

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

```{r}
install.packages("tinytable")
```

```{r}
#| include: false
#| warning: false
#| message: false

library(boot)
library(janitor)
library(modelsummary)
library(rstanarm)
library(tidyverse)
library(collapse)
library(tinytable)
library(marginaleffects)
```

```{r}
# combine all the years data for each ward and for each category
model_data <- analysis_budget_data |>
  group_by(ward_id, category, ward) |>
  summarize(total_10_year = sum(total_10_year)) |>
  pivot_wider(names_from = category, values_from = total_10_year)

# combine the model data with the ward profile data
model_data <- model_data |>
  left_join(analysis_ward_data, by = "ward_id") |>
  select(ward_id, ward.x, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
  rename(ward = ward.x) |>
  left_join(analysis_building_permits, by = "ward_id") |>
  rename(growth_related = `Growth Related`, state_of_good_repair = `State of Good Repair`, service_improvement = `Service Improvement and Enhancement`)

```

```{r}
# Install and load rstanarm if not installed
# install.packages("rstanarm")
# Load required libraries
library(rstanarm)
library(bayesplot)

# Prepare the data (assuming you've read the CSV)
data <- read.csv(here::here("data/02-analysis_data/model_data.csv"))

write.csv(data, here::here("data/02-analysis_data/model_data.csv"))

# Basic Poisson Model
permits_model <- stan_glm(
  building_permits ~ ward_id,
  data = model_data,
  family = poisson(link = "log"),s
  prior = normal(0, 2.5),  # Weakly informative prior
  prior_intercept = normal(0, 10),
  chains = 4,
  iter = 2000,
  seed = 123
)

# Model summary
summary(permits_model)

# Diagnostic plots
plot(permits_model)

# Posterior predictive checks
pp_check(permits_model)

# Credible intervals
posterior_interval(permits_model)

# Visualize posterior distributions
mcmc_areas(permits_model)

# Predictive performance
# Compare observed vs. predicted
predicted_permits <- posterior_predict(permits_model)

```

```{r}
library(rstanarm)
library(bayesplot)

# Fit the Negative Binomial model
# building_permits is the target variable
# population, average_household_income, growth_related, state_of_good_repair, and service_improvement are predictors
model <- stan_glmer(
  growth_related ~ population + average_household_income + (1 | ward_id),
  data = data,
  family = neg_binomial_2(link = "log"),
  chains = 4,
  iter = 2000,
  seed = 1234
)

# Summarize the model
summary(model)

# Posterior predictive checks
pp_check(model)

# Diagnostic plots
plot(model)

# Posterior predictive checks
pp_check(model)

# Credible intervals
posterior_interval(model)

# Visualize posterior distributions
mcmc_areas(model)

# Predictive performance
# Compare observed vs. predicted
predicted_permits <- posterior_predict(model)
```

```{r}
# model summary

```

### 2024-12-02 paper/paper.qmd
Initial Commit
Initial Commit

---
title: "Who Gets What: How Population and Income Shape Toronto’s Budget Decisions"
subtitle: "Understanding the Impact of Demographics and Economic Factors on City Spending Across 25 Wards"
author: 
  - Aviral Bhardwaj
thanks: "Code and data are available at: https://github.com/Aviral-03/Toronto-WardWide-Budget-Analysis"
date: today
date-format: long
abstract: "This study investigates how population density and average household income affect budget allocations across Toronto’s 25 wards. Using data from the city's 2023-2032 Capital Budget Plan and 2021 Ward Profiles, we analyze spending patterns in key areas like health and safety, infrastructure, and community services. Our findings reveal that higher population densities do not guarantee proportionally greater investments in essential services, while wealthier wards often receive larger capital expenditures. These insights highlight potential disparities in resource distribution, emphasizing the need for more equitable urban budget planning to address diverse community needs."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
prefer-html: true
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(opendatatoronto)
library(readxl)
library(styler)
library(dplyr)
library(arrow)
library(ggrepel)
library(knitr)
library(kableExtra)
library(patchwork)
library(sf)
library(lintr)
library(DiagrammeR)
library(rsvg)
library(magrittr)
library(DiagrammeRsvg)
library(png)
library(webshot2)
library(gridExtra)
```

```{r}
#| include: false
#| warning: false
#| message: false

# Get the analysis data
analysis_budget_data <- read.csv(here::here("data/02-analysis_data/analysis_budget_data.csv"))
analysis_building_permits <- read.csv(here::here("data/02-analysis_data/analysis_building_permits.csv"))
analysis_ward_data <- read.csv(here::here("data/02-analysis_data/analysis_ward_data.csv"))

# Get Model
# model <- readRDS(here::here("models/model.rds"))
```

# Introduction {#sec-intro}

# Data {#sec-data}

The raw data was sourced from the City of Toronto's Open Data Portal using the `opendatatoronto` [@openDataToronto] package. For the purpose of this analysis we used several data sets: `2023 Ward Profiles (25-Ward Model)` [@toronto_ward_profiles], `Capital Budget and Plan Details from 2021-2024` [@toronto_budget_plan], `City Wards` [@toronto_city_wards], and `Building Permits - Active Permits` [@toronto_building_permits].

The data, provided in CSV formats, was cleaned and analyzed using R [@citeR] programming language. The `readxl` [@readxl] package was used for reading Excel files. Other R packages used which includes `tidyverse` [@tidyverse], `styler` [@styler], and `dplyr` [@dplyr] for creating tables. The `ggplot2` [@ggplot2] and `kableExtra` [@kableExtra] were used for data visualization and table formatting. The `patchwork` [@patchwork] package was used for combining multiple plots, and `sf` [@sf] for spatial data analysis. For models, we used `rstan` [@rstan] was used for fitting the model, and `modelsummary` [@modelsummary] for model summary tables. The `lintr` [@lintr] package was used for code linting. The `arrow` [@arrow] package was used for reading and writing Parquet files.

## Measurement

Our research question and estimand analyse the relationship between key demographic and economic factors—specifically, population and average household income—and the budget allocation across various categories in Toronto's 25 wards. Population is a critical demographic indicator, representing the number of residents in each ward, while average household income reflects economic well-being, influencing access to resources and quality of life [@Schaeffer]. Therefore, these two variables of interest were chosen for the analysis. Our primary aim is to estimate the impact of these variables on budget allocations for these specific categories of interest, specifically Growth-Related expenditures, State of Good Repair, and Service Improvement and Enhancement.

Population density is expected to affect the demand for services and infrastructure, while average household income may shape how resources are distributed across wards, reflecting broader economic disparities. The estimand is the causal effect of these factors on the allocation of funds, and the estimator will quantify how shifts in population and income levels influence budgetary decisions. By understanding this relationship, we aim to offer insights into how demographic and economic variables drive municipal spending patterns in different categories across the city. In order to better analyze the trends and modelling the data, we will use the Captial Budget and Plan Details from 2021-2024 datasets, comparing total 10-year budget allocations across different wards and categories of interest.

## Ward Profiles (25-Ward Model) {#sec-ward-profiles}

The 2021 Ward Profiles [@toronto_ward_profiles], based on the 25-Ward model were provided by City Planning. These profiles included census data from the 2021, 2016, and 2011 Census of Population, covering demographic, social, and economic information for each ward in Toronto. These variables were collected through methods including online responses, mailed questionnaires, the Census Help Line, and enumerators [@nhs_canada].

These questionnaires gathered information on various topics related to residents' demographic characteristics, such as education, household income, number of dependents, employment status and etc. Participation in the survey is voluntary, and data is collected directly from residents, including their postal codes, which are used to determine their respective wards. To ensure privacy and confidentiality, the data is subsequently aggregated and anonymized [@canada_statcan_2023_nhs].

This data-set was included in this analysis to provide insights into the population and average household income for each ward, providing insights into the city's socioeconomic landscape. 25-Ward model was used instead of the 44-Ward model as it was the most recent data available at the time of analysis and matched the Capital Budget data.

The data was stored in an Excel workbook with multiple tabs, but for this analysis, we used the first tab, `2021 Census One variable`, which contains data for all 25 wards (Ward 1, Ward 2, ..., Ward 25). After cleaning, the data was saved in CSV and Parquet formats, with the following columns:

-   `ward_id`: unique identifier for each ward,
-   `ward`: ward name,
-   `population`: total population,
-   `income`: average household income.

The ward names were manually entered into the cleaned data to match with `ward_id`. A sample of the data can be seen in [@tbl-ward-profile-data].

```{r}
#| echo: false
#| message: false
#| label: tbl-ward-profile-data
#| tbl-cap: Sample of Cleaned Toronto Ward Profile Data
#| tbl-pos: "h"

# Display the first few rows of the cleaned ward profile data using kable
analysis_ward_data |>
  head() |>
  kable(col.names = c("Ward ID", "Ward Name", "Population", "Income")) |>
  kable_styling(full_width = F, position = "center")

```

## Capital Budget and Plan Details {#sec-capital-budget}

Each year, the City of Toronto publishes the Capital Budget and Plan Details dataset [@toronto_budget_plan], which outlines a 10-year capital budget and plan. This dataset breaks down the capital budget across the city’s 25 wards, allocating funds for infrastructure projects, equipment purchases, and other fixed assets. This budget is developed through a collaborative process, where city staff prepare an initial draft, which is then reviewed by the Budget Committee. Input is solicited from Toronto residents and businesses, and subsequently, the Mayor presents the finalized budget proposal by February 1. City Council reviews and considers this budget within 30 days [@toronto_2024_budget].

For the purpose of this analysis we selected the year 2021-2024 to align with the 2022 municipal elections and the subsequent relevance to planning efforts. Furthermore, the city's budgeting process underwent significant shifts after 2020 due to the impact of the COVID-19 pandemic, which altered spending priorities and resource allocation. Focusing on the 2021–2024 timeframe allows us to analyze the post-pandemic period, avoiding the uncertainties of the pandemic and ensuring the data remains consistent and reliable.

Each budget plan includes five primary categories under `State of Good Repair`, `Growth Related`, `Health and Safety`, `Service Improvement and Enhancement`, and `Legislated`. These categories define the main areas where capital expenditures are directed. For this analysis, however, we will focus on these three variables of interest: `State of Good Repair`, `Growth Related`, and `Service Improvement and Enhancement`.

These categories were selected because they represent critical areas of investment that directly impact the quality of life and well-being of Toronto residents. `State of Good Repair` focuses on maintaining and preserving existing infrastructure, `Growth Related` addresses the expansion and development of new infrastructure, and `Service Improvement and Enhancement` aims to enhance public services and amenities. By analyzing budget allocations in these categories, we can gain valuable insights into the city's spending priorities and identify opportunities for more effective and equitable resource distribution. While also identifying potential disparities in funding across various wards of the city.

Raw data includes key columns such as `Project Name`, yearly budget allocations for each year, `Ward Number`, `Ward`, `Category`, and `Total 10 Year` (Sum of Year 1 to 10), where the budget is in thousands of dollars (e.g., 10 = \$10,000). [@tbl-capital-budget-data] shows a sample of the cleaned data along with our variables of interest:

-   `Ward ID`
-   `Ward Name`
-   `Category of the Capital Budget`
-   `Total 10-year capital budget` allocated to each ward

Rows with `CW` (city-wide budget) were removed since they were applicable to all wards.

```{r}
#| echo: false
#| message: false
#| label: tbl-capital-budget-data
#| tbl-cap: Sample of Cleaned Toronto Capital Budget Data
#| tbl-pos: "h"

analysis_budget_data |>
  head() |>
  select(ward_id, ward, category, total_10_year) |>
  kable(col.names = c("Ward ID", "Ward Name", "Category", 
                  "Total 10-Year Budget (in 000s)"))
```

## City Wards {#sec-ward-names}

The City Wards dataset [@toronto_city_wards], published by the City Clerk's Office and last updated on July 22, 2024, contains geographical information about each ward, including the ward ID, ward name, and ward boundary. These ward boundaries were decided as a part of `Bill 5, Better Local Government Act` in 2018, reducing the number of wards from 47 to 25 [@toronto_city_wards].

This dataset, effective January 1, 2024, was used to map the `ward_id` to the ward name in the cleaned data. Key columns include:

-   `ward_id`: unique identifier for each ward,
-   `ward`: ward name,
-   `ward_boundary`: geographical boundary of the ward.

The ward names were mapped to the `ward_id` and integrated with the Ward Profiles [@sec-ward-profiles] and Capital Budget data sets [@sec-capital-budget] to create the final data set for analysis. This dataset was not used directly in the analysis but was essential for mapping the ward names to the ward IDs in the cleaned data.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-toronto-map-population
#| fig-cap: "Map of Toronto highlighting the population and income densities by ward"
#| fig.pos: 'H'
#| fig.height: 5

# URL to the zip file
url <- "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/5e7a8234-f805-43ac-820f-03d7c360b588/resource/35f67d86-cfc8-4483-8d77-50d035b010d9/download/25-ward-model-december-2018-wgs84-latitude-longitude.zip"

# Temporary file to store the downloaded zip
temp_zip <- tempfile(fileext = ".zip")

# Download the zip file
download.file(url, temp_zip, mode = "wb")

# Unzip the file to a temporary directory
temp_dir <- tempdir()
unzip(temp_zip, exdir = temp_dir)

shapefiles <- list.files(temp_dir, pattern = "\\.shp$", full.names = TRUE)

toronto_map <- st_read(shapefiles[1], quiet = TRUE)

# Clean up the downloaded zip file
unlink(temp_zip)

# Merge analysis_data with toronto_map by ward_id
toronto_map <- merge(toronto_map, analysis_ward_data, 
                     by.x = "AREA_NAME", by.y = "ward") |>
  rename("income" = "average_household_income")

color_scale <- scale_fill_gradient(low = "lightblue", high = "darkblue")

# Plot with facet_wrap for average household income and population
plot1 <-ggplot() +
  geom_sf(data = toronto_map, aes(fill = population)) +
  geom_sf_text(data = toronto_map, aes(label = ward_id), 
               size = 2, color = "white") +
  color_scale +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude") +
  theme(legend.position = "right")

plot2 <- ggplot() +
  geom_sf(data = toronto_map, aes(fill = income)) +
  geom_sf_text(data = toronto_map, aes(label = ward_id), size = 2, color = "white") +
  color_scale +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude") +
  theme(legend.position = "right")

plot1 / plot2

```

## Building Permits - Active Permits {#sec-building-permits}

The *Building Permits - Active Permits* dataset [@toronto_building_permits], published by the City Planning Division and last updated on November 28, 2024, serves as a comprehensive record of active building permits in Toronto. A building permit is a municipally issued document, mandated by the Building Code Act and enforced by the City of Toronto, regulating the construction or demolition of physical structures [@toronto_building_permits].

The process of obtaining a building permit involves submitting an application to the City of Toronto, including necessary drawings, documents, and other forms based on the permit type. The Building Division reviews the application, and a Toronto Building Inspector ensures compliance with the Ontario Building Code, Zoning By-law, and other applicable regulations. Once approved, the permit is issued, allowing the applicant to commence construction or demolition.

Table [@tbl-raw-building-permits-data] outlines detailed information about active building permits in Toronto, including key features such as `Permit Number`, `Permit Type`, `Structure Type`, `Status`, and more.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-raw-building-permits-data
#| tbl-cap: "Sample of the raw building permits data"
#| tbl-pos: "h"

# Load the raw building permits data
building_permits <- read_csv(here::here("data/01-raw_data/building_permits_data.csv"))

# Display the first few rows of the building permits data using kable
building_permits |>
  select("_id", PERMIT_TYPE, STRUCTURE_TYPE, WORK, POSTAL, STATUS) |>
  head() |>
  kable(col.names = c("ID", "Permit Type", "Structure Type", "Work", "Postal Code", "Status"))
```

We selected this data-set to evaluate how budget allocations correlate with the number of building permits issued in each ward. The number of permits serves as a proxy for construction activity and development, highlighting the demand for infrastructure investment and capital expenditures. By examining the relationship between building permits and budget allocations, we can gain valuable insights into how construction activity influences resource distribution across the city's wards.

From the raw data, we selected key variables of interest: `Postal Code`, `Status`, and `Work`. The `Postal Code` was used to map building permits to their respective wards, while `Status` and `Work` were utilized to filter permits based on their current state and type of work. [@tbl-status-building-permits-data] lists 45 unique statuses assigned to each submitted permit. For this analysis, we focused on permits with the following statuses: `Approved`, `Application Accepted`, `Issuance Pending`, `Ready for Issuance`, `Permit Issued`, and `Permit Issued/Close File`. These statuses indicate that the permit has been approved and that construction or demolition is either underway or completed.

To ensure we analyzed only construction-related permits, we filtered out entries where `Work` did not have the value `New Building`. The cleaned data was then saved in CSV and Parquet formats with the following columns:

-   **`ward_id`**: Unique identifier for each ward.\
-   **`total_building_permits`**: Total number of building permits issued in each ward.

[@fig-building-permits-data] illustrates the total number of building permits by ward, providing valuable insights into construction activity and development patterns across Toronto’s wards.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-building-permits-data
#| fig-cap: "Total number of building permits by Ward"
#| fig-pos: "h"
#| fig-height: 5
#| fig-width: 10

# Display the total number of building permits by ward using kable
analysis_building_permits |>
  ggplot(aes(x = ward_id, y = total_building_permits)) +
  geom_histogram(stat = "identity", fill = "skyblue", color = "darkblue") +
  labs(x = "Ward ID", y = "Number of building permits") +
  scale_x_continuous(breaks = seq(1, 25, 1)) +
  # change the theme
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

## Trends in Budget Allocations (2021-2024) {#sec-budget-trends}

## Model {#sec-model}

The purpose of this paper is to investigate the relationship between population density, average household income, and budget allocations across Toronto's 25 wards, and their potential impact on the economic well-being of each ward, as represented by the `total_building_permits` issued. To achieve this, we developed a causal model that examines how population and income influence the allocation of funds in key budget categories: `Growth Related`, `State of Good Repair`, and `Service Improvement and Enhancement`.

We employed multilevel modeling to estimate the effect of these variables on the number of building permits issued in each ward. Here we briefly describe a Bayesian Poisson regression model, which incorporates varying intercepts by grouping the effect by ward account. This approach allows us to capture the nuanced effects of demographic and economic factors on infrastructure development and resource distribution. Background details and diagnostics are included in [Appendix @sec-model-details].

### Model set-up

The model was specified as follows:

\begin{align*}
\text{Building Permits}_{i} & \sim \text{Poisson}(\lambda_{i}) \\
\log(\lambda_{i}) & = \beta_{0} + \beta_{1} \times \text{Population}_{i} + \beta_{2} \times \text{Income}_{i} + \beta_{3} \times \text{Budget}_{i} + \alpha_{\text{ward}[i]} \\
\beta_{0} & \sim \text{Normal}(0, 2.5) \\
\beta_{1} & \sim \text{Normal}(0, 2.5) \\
\beta_{2} & \sim \text{Normal}(0, 2.5) \\
\beta_{3} & \sim \text{Normal}(0, 2.5) \\
\alpha_{\text{ward}} & \sim \text{Normal}(0, \sigma_{\text{ward}})
\end{align*}

In the above model:

-   $\text{Building Permits}_i$ represents the number of building permits issued in ward `i`,
-   $\lambda_i$ is the expected count of permits for ward $i$.
-   $\log(\lambda_i)$ is the linear predictor, which includes:
    -   $\beta_0$: Intercept
    -   $\beta_1$: Effect of `population`
    -   $\beta_2$: Effect of `average_household_income`
    -   $\beta_3$: Effect of `budget_allocation`
    -   $\alpha_{\text{ward}[i]}$: Random effect for each ward (captures unobserved heterogeneity between wards).
-   Priors for $\beta_0$, $\beta_1$, $\beta_2$, and $\beta_3$ are Normal(0, 2.5), allowing for flexibility.

The model was fitted using the `stan_glmer` function from the `rstanarm` package [@citerstanarm] in R [@citeR]. The prior distributions for the coefficients were set to normal distributions with a mean of 0 and a standard deviation of 2.5. The model also included a prior for the intercept, which was set to a normal distribution with a mean of 0 and a standard deviation of 2.5. The varying intercepts for each ward were modeled using a normal distribution with a mean of 0 and a standard deviation of $\sigma_{\text{ward}}$.

### Model justification

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-causal-model
#| fig-cap: "Causal model for the relationship between population, income, and budget allocations"
#| fig-pos: "h"

# create a dag for the model
causal_model <- "digraph {
  graph []
  # Nodes
  node [shape = plaintext]
    A [label = 'Population']
    B [label = 'Budget Allocation']
    C [label = 'Active Construction Projects']
    D [label = 'Income']
{ rank = same A D; }
  # Edges for confounding variable
  edge []
    A -> {B C}
    D -> {B C}
    B -> C
}"

# render the causal model
exporting_image <- grViz(causal_model) |>
  export_svg() |>
  charToRaw() |>
  rsvg_png(here::here("other/sketches/causal_model.png"), width = 800, height = 600)

image <- readPNG(here::here("other/sketches/causal_model.png"))
par(mar = rep(2, 4))
plot.new()
rasterImage(image, 0, 0, 1, 1)  
```

## Results {#sec-results}

In this section, we visualize the relationships between population density, average household income, budget allocations with total number of building permits across Toronto's 25 wards. Additionally, we present our model results, highlighting how these variables influence the number of building permits issued in each ward.

### Relationship between Population, Average Household Income, and Budget Allocations

Our budget data spans three categories—`Growth Related`, `State of Good Repair`, and `Service Improvement and Enhancement`—for the years 2021-2024. As observed in [@sec-budget-trends], there has been a steady increase in budget allocations across these categories. To better understand the relationship between these budget categories and our variables of interest, we calculate the average budget allocation for each category across the years. The results are presented in [@tbl-avg-budget-data].

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-avg-budget-data
#| tbl-cap: "Average 10-year budget allocations by category (2021-24)"
#| tbl-pos: "h"

# Calculate the average 10-year budget allocations by category
average_budget <- analysis_budget_data |>
  group_by(ward_id, ward, category) |>
  summarise(avg_total_10_year = mean(total_10_year)) |>
  pivot_wider(names_from = category, values_from = avg_total_10_year) |>
  select(ward_id, ward, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`)

average_budget |>
  head() |>
  kable(col.names = c("Ward ID", "Ward Name", "Growth Related", 
                      "State of Good Repair", "Service Improvement and Enhancement"))
```
## Relationship between Average Household Income & Population with Budget Allocation for Growth Related Projects

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-growth-related-budget
#| fig.cap: "Growth Related Budget by Ward"
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 5

combined_data <- analysis_ward_data |>
  left_join(average_budget, by = "ward_id") |>
  rename(ward = ward.x) |>
  select(ward_id, ward, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`)

growth_budget <- combined_data |>
  select(ward_id, ward, population, average_household_income, `Growth Related`)

p1 <- ggplot(growth_budget, aes(x = population, y = `Growth Related`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
  theme_minimal()

p2 <- ggplot(growth_budget, aes(x = average_household_income, y = `Growth Related`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Average Household Income", y = "") +
  theme_minimal()

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)

```

## Relationship between Average Household Income & Population with Budget Allocation for State of Good Repair Projects

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-state-of-good-repair-budget
#| fig.cap: "State of Good Repair Budget by Ward"
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 5

good_repair_budget <- combined_data |>
  select(ward_id, ward, population, average_household_income, `State of Good Repair`)

p1 <- ggplot(good_repair_budget, aes(x = population, y = `State of Good Repair`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
  theme_minimal()

p2 <- ggplot(good_repair_budget, aes(x = average_household_income, y = `State of Good Repair`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Average Household Income", y = "") +
  theme_minimal()

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)
```

### Relationship between Average Household Income & Population with Budget Allocation for Service Improvement and Enhancement Projects
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-service-improvement-budget
#| fig.cap: "Service Improvement and Enhancement Budget by Ward"
#| fig-pos: "h"
#| fig.width: 10
#| fig.height: 5

s_i_budget <- combined_data |>
  select(ward_id, ward, population, average_household_income, `Service Improvement and Enhancement`)

p1 <- ggplot(s_i_budget, aes(x = population, y = `Service Improvement and Enhancement`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
  theme_minimal()

p2 <- ggplot(s_i_budget, aes(x = average_household_income, y = `Service Improvement and Enhancement`, label= ward_id)) +
  geom_point() +
  geom_text_repel(hjust = -0.5, vjust = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(x = "Average Household Income", y = "") +
  theme_minimal()

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)
```

# Discussion {#sec-discussion}

\newpage

# Appendix {#sec-appendix}

## Data Tables

### Building Permits Status

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-status-building-permits-data
#| tbl-cap: "Building permit statuses"
#| tbl-pos: H

# Display the unique statuses in the building permits data using kable
building_permits |>
  distinct(STATUS) |>
  kable() |>
  kable_styling(full_width = F, position = "center")

```

## Model Details {#sec-model-details}

```{r}
#| echo: false
#| warning: false
#| message: false
# select only growth related, state of good repair and service improvement and enhancement
service_improvement <- analysis_budget_data |>
  filter(category == "Service Improvement and Enhancement") |>
  mutate(total_10_year = total_10_year / 1000) |>
  # remove negative values
  filter(total_10_year > 0) |>
 rename(total_budget = total_10_year)

# Calculate the outliers using IQR
outliers <- service_improvement |>
  group_by(year) |>
  mutate(
    Q1 = quantile(total_budget, 0.25),
    Q3 = quantile(total_budget, 0.75),
    IQR = Q3 - Q1,
    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
  ) |>
  filter(is_outlier)

# Plot with labels for outliers
ggplot(service_improvement, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    title = "Total Budget Over Years",
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| warning: false
#| message: false
# select only growth related, state of good repair and service improvement and enhancement
growth_related <- analysis_budget_data |>
  filter(category == "Growth Related") |>
  mutate(total_10_year = total_10_year / 1000) |>
  # remove negative values
  filter(total_10_year > 0) |>
 rename(total_budget = total_10_year)

# Calculate the outliers using IQR
outliers <- growth_related |>
  group_by(year) |>
  mutate(
    Q1 = quantile(total_budget, 0.25),
    Q3 = quantile(total_budget, 0.75),
    IQR = Q3 - Q1,
    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
  ) |>
  filter(is_outlier)

# Plot with labels for outliers
ggplot(growth_related, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    title = "Total Budget Over Years",
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| warning: false
#| message: false
# select only growth related, state of good repair and service improvement and enhancement
state_of_good_repair <- analysis_budget_data |>
  filter(category == "State of Good Repair") |>
  mutate(total_10_year = total_10_year / 1000) |>
  # remove negative values
  filter(total_10_year > 0) |>
 rename(total_budget = total_10_year)

# Calculate the outliers using IQR
outliers <- state_of_good_repair |>
  group_by(year) |>
  mutate(
    Q1 = quantile(total_budget, 0.25),
    Q3 = quantile(total_budget, 0.75),
    IQR = Q3 - Q1,
    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
  ) |>
  filter(is_outlier)

# Plot with labels for outliers
ggplot(state_of_good_repair, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    title = "Total Budget Over Years",
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

# References

### 2024-12-02 scripts/02-download_data.R

@@ -1,0 +1,52 @@
+#### Preamble ####
+# Purpose: Downloads and saves the data from Open Data Toronto.
+# Author: Aviral Bhardwaj
+# Date: 2024-12-01
+# Contact: aviral.bhardwaj@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: N.A
+# Any other information needed? N.A
+
+
+#### Work-space setup ####
+
+library(tidyverse)
+library(readxl)
+library(styler)
+library(dplyr)
+library(here)
+
+#### Download data ####
+
+# Define a function to download, read, and save data
+download_and_save <- function(url, output_file) {
+  local_file <- tempfile(fileext = ".csv")
+  download.file(url, local_file, mode = "wb")
+  if (grepl("xlsx$", url)) {
+    data <- readxl::read_xlsx(local_file)
+  } else
+    data <- readr::read_csv(local_file)
+  write_csv(data, output_file)
+  unlink(local_file)
+  return(data)
+}
+
+# List of URLs and output file names
+datasets <- list(
+  "ward_profile_data" = "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/6678e1a6-d25f-4dff-b2b7-aa8f042bc2eb/resource/16a31e1d-b4d9-4cf0-b5b3-2e3937cb4121/download/2023-WardProfiles-2011-2021-CensusData.xlsx",
+  "budget_data_2024" = "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/7d3bcf2f-8eca-4ed5-a352-a34adb130931/resource/6b774b3a-5e1a-4362-ba31-a3b07fce31db/download/2024-2033-capital-budget-and-plan-details.xlsx",
+  "budget_data_2023" = "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/7d3bcf2f-8eca-4ed5-a352-a34adb130931/resource/50f76ab0-3ed3-41b4-8350-49c2c52911f9/download/2023-2032-capital-budget-and-plan-details.xlsx",
+  "budget_data_2022" = "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/7d3bcf2f-8eca-4ed5-a352-a34adb130931/resource/2640ca61-9eae-4d0d-89b2-523952c2b611/download/2022-2031-capital-budget-and-plan-details.xlsx",
[…]

### 2024-12-02 scripts/03-clean_data.R

@@ -1,0 +1,209 @@
+#### Preamble ####
+# Purpose: Cleans the raw ward profile, ward capital budget data and building permit data
+# Author: Aviral Bhardwaj
+# Date: 2024-12-01
+# Contact: aviral.bhardwaj@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: N.A
+# Any other information needed? N.A
+
+#### Works-pace setup ####
+library(tidyverse)
+library(here)
+
+#### Clean data ####
+input_folder <- here::here("data/01-raw_data")
+output_folder <- here::here("data/02-analysis_data")
+
+## Cleaning Budget Data
+
+years <- c(2021, 2022, 2023)
+
+# Function to clean data
+clean_capital_data <- function(file_path, year) {
+  raw_data <- read_csv(file_path)
+
+  cleaned_data <- raw_data[c("Ward Number", "Ward", "Category", "Total 10 Year")] |>
+    rename(
+      ward_id = "Ward Number",
+      ward = "Ward",
+      category = "Category",
+      total_10_year = "Total 10 Year"
+    ) |>
+    filter(ward_id != "CW") |>
+    group_by(ward_id, ward, category) |>
+    summarise(total_10_year = sum(total_10_year), .groups = "drop") |>
+    mutate(ward_id = as.numeric(ward_id)) |>
+    arrange(ward_id)
+
+  # Save cleaned data to a csv file
+  output_file <- file.path(output_folder, paste0("cleaned_data_", year, ".csv"))
+  write.csv(cleaned_data, output_file, row.names = FALSE)
+
+  return(cleaned_data)
+}
+
+# Process and clean data for each year
+cleaned_data <-
+  lapply(years, function(year) {
+    file_path <- file.path(input_folder, paste0("budget_data_", year, ".csv"))
+    clean_capital_data(file_path, year)
+  })
+
+
+cleaned_2024 <- read_csv(here::here("data/01-raw_data/budget_data_2024.csv"))
+
+# Only keep the relevant columns with name Total 10 Year, Ward Number, Ward, Category
+cleaned_capital_data <- cleaned_2024[c(14, 15, 16, 17)] |>
+  rename("ward_id" = "Ward Number") |>
+  rename("ward" = "Ward") |>
+  rename("category" = "Category") |>
+  rename("total_10years" = "total_10years")
+
+# Filter out the rows in ward_id that are CW (city wide) as they are not relevant
+cleaned_capital_data <- cleaned_capital_data %>%
[…]

### 2024-12-02 scripts/06-model_data.R

@@ -1,0 +1,48 @@
+#### Preamble ####
+# Purpose: Models... [...UPDATE THIS...]
+# Author: Rohan Alexander [...UPDATE THIS...]
+# Date: 11 February 2023 [...UPDATE THIS...]
+# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# License: MIT
+# Pre-requisites: [...UPDATE THIS...]
+# Any other information needed? [...UPDATE THIS...]
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(rstanarm)
+library(bayesplot)
+library(here)
+
+#### Read data ####
+analysis_budget_data <- read.csv(here::here("data/02-analysis_data/analysis_budget_data.csv"))
+analysis_building_permits <- read.csv(here::here("data/02-analysis_data/analysis_building_permits.csv"))
+analysis_ward_data <- read.csv(here::here("data/02-analysis_data/analysis_ward_data.csv"))
+
+### Model data ####
+
+# combine all the years data for each ward and for each category
+model_data <- analysis_budget_data |>
+  group_by(ward_id, category, ward) |>
+  summarize(total_10_year = sum(total_10_year)) |>
+  pivot_wider(names_from = category, values_from = total_10_year)
+
+# combine the model data with the ward profile data
+model_data <- model_data |>
+  left_join(analysis_ward_data, by = "ward_id") |>
+  select(ward_id, ward.x, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
+  rename(ward = ward.x) |>
+  left_join(analysis_building_permits, by = "ward_id") |>
+  rename(growth_related = `Growth Related`, state_of_good_repair = `State of Good Repair`, service_improvement = `Service Improvement and Enhancement`)
+
+
+#### Model ####
+
+
+#### Save model ####
+saveRDS(
+  first_model,
+  file = "models/first_model.rds"
+)
+
+

### 2024-12-03 paper/dump.qmd
Added New Test Scripts
- Added Simulation
- Added Simulation Tests
- Added Analysis Data Tests
- Added Sketches

@@
 ---
-
-## Park and Recreation Facility Projects {#sec-parks}
-
-The Park and Recreation Facility Projects dataset [@toronto_parks], published by the Parks, Forestry and Recreation Division and last updated on July 22, 2024, contains information about park and recreation facility projects in Toronto.
-This dataset includes details about the project name, location, ward number, and project status.
-Park and recreation facilities are essential public amenities that contribute to the quality of life and well-being of residents, providing spaces for recreation, leisure, and community engagement.
-
-We choose this dataset to evaluate the how budget allocations are related to the number of park and recreation facility projects in each ward.
-The number of park projects can serve as a proxy for community development and public service enhancement, reflecting the need for investments in public amenities and infrastructure.
-By analyzing the relationship between park projects and budget allocations, we can gain insights into how public service enhancements influence the distribution of resources across the city.
-This will also help us understand how the city's budget decisions align with community development trends and public service needs in different wards, and determine if there are any disparities in resource allocation based on public service enhancements to certain wards.
-
-```{r}
-parks_data <- here::here("data/01-raw_data/Park and Recreation Facility Projects - 4326.csv")
-parks <- read_csv(parks_data)
-
-# create a graph of the number of parks by ward_number
-parks |>
-  count(ward_number) |>
-  ggplot(aes(x = ward_number, y = n)) +
-  geom_col() +
-  labs(x = "Ward number", y = "Number of parks") +
-  theme_minimal()
-```
-
-```{r}
-#| include: false
-#| warning: false
-#| message: false
-
-library(tidyverse)
-library(palmerpenguins)
-```
 
@@
 ## Outcome variables
-
-Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.
-
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
-
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
-
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
-
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))
-
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
-```
-
-Talk way more about it.
 
@@
 library(marginaleffects)
+library(rstanarm)
+library(bayesplot)
 ```
@@
 ```{r}
+analysis_budget_data <- read.csv(here::here("data/02-analysis_data/analysis_budget_data.csv"))
+analysis_building_permits <- read.csv(here::here("data/02-analysis_data/analysis_building_permits.csv"))
+analysis_ward_data <- read.csv(here::here("data/02-analysis_data/analysis_ward_data.csv"))
+
 # combine all the years data for each ward and for each category
@@
   group_by(ward_id, category, ward) |>
-  summarize(total_10_year = sum(total_10_year)) |>
-  pivot_wider(names_from = category, values_from = total_10_year)
+  summarise(avg_total_10_year = mean(total_10_year)) |>
+  pivot_wider(names_from = category, values_from = avg_total_10_year) |>
+  # sum the total 10 year budget for each ward
+  group_by(ward_id, ward) |>
+  summarise(total_budget = sum(`Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`))
 
@@
   left_join(analysis_ward_data, by = "ward_id") |>
-  select(ward_id, ward.x, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
+  select(ward_id, ward.x, total_budget, average_household_income, population) |>
   rename(ward = ward.x) |>
+  # left join with the building permits data
   left_join(analysis_building_permits, by = "ward_id") |>
-  rename(growth_related = `Growth Related`, state_of_good_repair = `State of Good Repair`, service_improvement = `Service Improvement and Enhancement`)
-
+  select(ward_id, ward, total_budget, average_household_income, population, total_building_permits)
 ```
@@
 ```{r}
-# Install and load rstanarm if not installed
-# install.packages("rstanarm")
-# Load required libraries
-library(rstanarm)
-library(bayesplot)
 
-# Prepare the data (assuming you've read the CSV)
-data <- read.csv(here::here("data/02-analysis_data/model_data.csv"))
+# convert the ward_id to a factor
+model_data$ward_id <- as.factor(model_data$ward_id)
 
-write.csv(data, here::here("data/02-analysis_data/model_data.csv"))
-
-# Basic Poisson Model
-permits_model <- stan_glm(
-  building_permits ~ ward_id,
+model <- stan_glm(
+  formula = total_building_permits ~ total_budget + average_household_income + population + ward_id,
   data = model_data,
-  family = poisson(link = "log"),s
-  prior = normal(0, 2.5),  # Weakly informative prior
-  prior_intercept = normal(0, 10),
-  chains = 4,
-  iter = 2000,
-  seed = 123
+  family = gaussian(link = "identity"),
+  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE)
 )
 
-# Model summary
-summary(permits_model)
-
-# Diagnostic plots
-plot(permits_model)
-
-# Posterior predictive checks
-pp_check(permits_model)
-
-# Credible intervals
-posterior_interval(permits_model)
-
-# Visualize posterior distributions
-mcmc_areas(permits_model)
-
-# Predictive performance
-# Compare observed vs. predicted
-predicted_permits <- posterior_predict(permits_model)
-
-```
-
-```{r}
-library(rstanarm)
-library(bayesplot)
-
-# Fit the Negative Binomial model
-# building_permits is the target variable
-# population, average_household_income, growth_related, state_of_good_repair, and service_improvement are predictors
-model <- stan_glmer(
-  growth_related ~ population + average_household_income + (1 | ward_id),
-  data = data,
-  family = neg_binomial_2(link = "log"),
-  chains = 4,
-  iter = 2000,
-  seed = 1234
-)
-
-# Summarize the model
+# Summarize model output
 summary(model)
@@
 
-# Diagnostic plots
+# Plot the coefficient estimates
 plot(model)
-
-# Posterior predictive checks
-pp_check(model)
 
@@
 predicted_permits <- posterior_predict(model)
+
+# plot diagnostics
+plot(model, "trace")
+plot(model, "rhat")
+
+modelsummary(
+  list(
+    "Model" = model
+  ),
+)
 ```
@@
 ```{r}
-# model summary
 

### 2024-12-03 paper/paper.qmd
Added New Test Scripts
- Added Simulation
- Added Simulation Tests
- Added Analysis Data Tests
- Added Sketches

@@
 
-## Model {#sec-model}
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+
+# select only growth related
+growth_related <- analysis_budget_data |>
+  filter(category == "Growth Related") |>
+  mutate(total_10_year = total_10_year / 1000) |>
+  # remove negative values
+  filter(total_10_year > 0) |>
+ rename(total_budget = total_10_year)
+
+# Calculate the outliers using IQR
+outliers <- growth_related |>
+  group_by(year) |>
+  mutate(
+    Q1 = quantile(total_budget, 0.25),
+    Q3 = quantile(total_budget, 0.75),
+    IQR = Q3 - Q1,
+    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
+  ) |>
+  filter(is_outlier)
+
+# Plot with labels for outliers
+plot1 <- ggplot(growth_related, aes(x = factor(year), y = total_budget)) +
+  geom_boxplot(fill = "skyblue", color = "darkblue") +
+  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
+  geom_text(
+    data = outliers, 
+    aes(label = ward), 
+    nudge_y = 15, 
+    check_overlap = TRUE
+  ) +  # Label only outliers
+  labs(
+    x = "Year",
+    y = "Total Budget (in thousands)"
+  ) +
+  theme_minimal() +
+  # remove legend
+  theme(legend.position = "none")
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+
+# select only growth related, state of good repair and service improvement and enhancement
+state_of_good_repair <- analysis_budget_data |>
+  filter(category == "State of Good Repair") |>
+  mutate(total_10_year = total_10_year / 1000) |>
+  # remove negative values
+  filter(total_10_year > 0) |>
+ rename(total_budget = total_10_year)
+
+# Calculate the outliers using IQR
+outliers <- state_of_good_repair |>
+  group_by(year) |>
+  mutate(
+    Q1 = quantile(total_budget, 0.25),
+    Q3 = quantile(total_budget, 0.75),
+    IQR = Q3 - Q1,
+    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
+  ) |>
+  filter(is_outlier)
+
+# Plot with labels for outliers
+plot2 <- ggplot(state_of_good_repair, aes(x = factor(year), y = total_budget)) +
+  geom_boxplot(fill = "skyblue", color = "darkblue") +
+  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
+  geom_text(
+    data = outliers, 
+    aes(label = ward), 
+    nudge_y = 15, 
+    check_overlap = TRUE
+  ) +  # Label only outliers
+  labs(
+    x = "Year",
+    y = "Total Budget (in thousands)"
+  ) +
+  theme_minimal() +
+  # remove legend
+  theme(legend.position = "none")
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+# select only growth related, state of good repair and service improvement and enhancement
+service_improvement <- analysis_budget_data |>
+  filter(category == "Service Improvement and Enhancement") |>
+  mutate(total_10_year = total_10_year / 1000) |>
+  # remove negative values
+  filter(total_10_year > 0) |>
+ rename(total_budget = total_10_year)
+
+# Calculate the outliers using IQR
+outliers <- service_improvement |>
+  group_by(year) |>
+  mutate(
+    Q1 = quantile(total_budget, 0.25),
+    Q3 = quantile(total_budget, 0.75),
+    IQR = Q3 - Q1,
+    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
+  ) |>
+  filter(is_outlier)
+
+# Plot with labels for outliers
+plot3 <- ggplot(service_improvement, aes(x = factor(year), y = total_budget)) +
+  geom_boxplot(fill = "skyblue", color = "darkblue") +
+  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
+  geom_text(
+    data = outliers, 
+    aes(label = ward), 
+    nudge_y = 15, 
+    check_overlap = TRUE
+  ) +  # Label only outliers
+  labs(
+    x = "Year",
+    y = "Total Budget (in thousands)"
+  ) +
+  theme_minimal() +
+  # remove legend
+  theme(legend.position = "none")
+```
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-budget-trends
+#| fig.cap: "Trends in budget allocations (2021-2024) by category"
+#| fig-subcap: ["Growth Related", "State of Good Repair", "Service Improvement and Enhancement"]
+#| fig-pos: "h"
+#| layout-ncol: 2
+
+plot1
+plot2
+plot3
+```
+
+## Relationship between Population, Average Household Income, and Budget Allocations
+
+Our budget data spans three categories—`Growth Related`, `State of Good Repair`, and `Service Improvement and Enhancement`—for the years 2021-2024.
+As observed in [@sec-budget-trends], there has been a steady increase in budget allocations across these categories.
+To better understand the relationship between these budget categories and our variables of interest, we calculate the average budget allocation for each category across the years.
+The results are presented in [@tbl-avg-budget-data].
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: tbl-avg-budget-data
+#| tbl-cap: "Average 10-year budget allocations by category (2021-24)"
+#| tbl-pos: "h"
+
+# Calculate the average 10-year budget allocations by category
+average_budget <- analysis_budget_data |>
+  group_by(ward_id, ward, category) |>
+  summarise(avg_total_10_year = mean(total_10_year)) |>
+  pivot_wider(names_from = category, values_from = avg_total_10_year) |>
+  select(ward_id, ward, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`)
+
+average_budget |>
+  head() |>
+  kable(col.names = c("Ward ID", "Ward Name", "Growth Related", 
+                      "State of Good Repair", "Service Improvement and Enhancement"))
+```
+### Relationship between Average Household Income & Population with Budget Allocation for Growth Related Projects and State of Good Repair Projects
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-growth-related-budget
+#| fig.cap: "Growth Related Budget by Ward"
+#| fig-pos: "h"
+#| fig.width: 10
+#| fig.height: 5
+
+combined_data <- analysis_ward_data |>
+  left_join(average_budget, by = "ward_id") |>
+  left_join(analysis_building_permits, by = "ward_id") |>
+  rename(ward = ward.x) |>
+  select(ward_id, ward, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`, total_building_permits)
+
+growth_budget <- combined_data |>
+  select(ward_id, ward, population, average_household_income, `Growth Related`)
+
+p1 <- ggplot(growth_budget, aes(x = population, y = `Growth Related`, label= ward_id)) +
+  geom_point() +
+  geom_text_repel(hjust = -0.5, vjust = 0.5) +
+  geom_smooth(method = "lm", color = "blue", se = TRUE) +
+  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
+  theme_minimal()
+
+p2 <- ggplot(growth_budget, aes(x = average_household_income, y = `Growth Related`, label= ward_id)) +
+  geom_point() +
+  geom_text_repel(hjust = -0.5, vjust = 0.5) +
+  geom_smooth(method = "lm", color = "blue", se = TRUE) +
+  labs(x = "Average Household Income", y = "") +
+  theme_minimal()
+
+# Combine both plots side by side
+grid.arrange(p1, p2, ncol = 2)
+
+```
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-state-of-good-repair-budget
+#| fig.cap: "State of Good Repair Budget by Ward"
+#| fig-pos: "h"
+#| fig.width: 10
+#| fig.height: 5
+
+good_repair_budget <- combined_data |>
+  select(ward_id, ward, population, average_household_income, `State of Good Repair`)
+
+p1 <- ggplot(good_repair_budget, aes(x = population, y = `State of Good Repair`, label= ward_id)) +
+  geom_point() +
+  geom_text_repel(hjust = -0.5, vjust = 0.5) +
+  geom_smooth(method = "lm", color = "blue", se = TRUE) +
+  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
+  theme_minimal()
+
+p2 <- ggplot(good_repair_budget, aes(x = average_household_income, y = `State of Good Repair`, label= ward_id)) +
+  geom_point() +
+  geom_text_repel(hjust = -0.5, vjust = 0.5) +
+  geom_smooth(method = "lm", color = "blue", se = TRUE) +
+  labs(x = "Average Household Income", y = "") +
+  theme_minimal()
+
+# Combine both plots side by side
+grid.arrange(p1, p2, ncol = 2)
+```
+[@fig-growth-related-budget] illustrates the relationship between average household income, population, and the total 10-year budget allocation for Growth-Related projects.
+The linear regression line for average household income remains relatively flat, hovering around \$100,000.
+This suggests that budget allocations for Growth-Related projects are not significantly influenced by income levels, indicating a more uniform distribution of funds across wards regardless of household income.
+A similar lack of correlation is observed in [@fig-state-of-good-repair-budget], where budget allocations for State of Good Repair projects also show minimal dependence on average household income.
+
+In contrast, the linear regression line for population shows a positive correlation with budget allocation for Growth-Related projects.
+Wards with higher population densities—such as Ward 10 (Spadina-Fort York), Ward 13 (Toronto Centre), Ward 5 (York South-Weston), and Ward 2 (Etobicoke North)—receive a larger share of the Growth-Related budget.
+This indicates that population density plays a critical role in determining funding levels for these projects, with more populous wards benefiting from increased allocations.
+A similar trend is observed in the budget distribution for State of Good Repair projects, where wards with higher population densities receive a proportionally larger share of the funds.
+
+### Relationship between Average Household Income & Population with Budget Allocation for Service Improvement and Enhancement Projects 
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-service-improvement-budget
+#| fig.cap: "Service Improvement and Enhancement Budget by Ward"
+#| fig-pos: "h"
+#| fig.width: 10
+#| fig.height: 5
+
+s_i_budget <- combined_data |>
+  select(ward_id, ward, population, average_household_income, `Service Improvement and Enhancement`)
+
+p1 <- ggplot(s_i_budget, aes(x = population, y = `Service Improvement and Enhancement`, label= ward_id)) +
+  geom_point() +
+  geom_text_repel(hjust = -0.5, vjust = 0.5) +
+  geom_smooth(method = "lm", color = "blue", se = TRUE) +
+  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
+  theme_minimal()
+
+p2 <- ggplot(s_i_budget, aes(x = average_household_income, y = `Service Improvement and Enhancement`, label= ward_id)) +
+  geom_point() +
+  geom_text_repel(hjust = -0.5, vjust = 0.5) +
+  geom_smooth(method = "lm", color = "blue", se = TRUE) +
+  labs(x = "Average Household Income", y = "") +
+  theme_minimal()
+
+# Combine both plots side by side
+grid.arrange(p1, p2, ncol = 2)
+```
+[@fig-service-improvement-budget] illustrates the relationship between average household income, population, and the total 10-year budget allocation for Service Improvement and Enhancement projects.
+These projects  aim to enhance the quality of services provided to residents and improve the overall infrastructure of the city [@toronto_improvements].
+
+Unlike Growth-Related and State of Good Repair projects, the linear regression lines for average household income and population shows a positive correlation with budget allocation for Service Improvement and Enhancement projects.
+Wards with higher average household incomes, such as Ward 11 (University-Rosedale) and Ward 13 (Toronto Centre), receive a larger share of the budget allocation for these projects.
+This suggests that higher-income areas are prioritized for service improvements and infrastructure enhancements, reflecting a targeted approach to resource allocation based on income levels.
+
+Other wards like Ward 3 (Etobicoke-Lakeshore) and Ward 10 (Spadina-Fort York), which have high population densities, also received significant budget allocations, suggesting that these areas are priorities for service improvements and infrastructure enhancement.
+
+Despite these outliers, the majority of wards received similar budget allocations for Service Improvement and Enhancement projects, underscoring a city-wide strategy to enhance services uniformly for all residents.
+
+# Model {#sec-model}
 
 The purpose of this paper is to investigate the relationship between population density, average household income, and budget allocations across Toronto's 25 wards, and their potential impact on the economic well-being of each ward, as represented by the `total_building_permits` issued.
-To achieve this, we developed a causal model that examines how population and income influence the allocation of funds in key budget categories: `Growth Related`, `State of Good Repair`, and `Service Improvement and Enhancement`.
+To achieve this, we will first plot the data to visualize the relationships between these variables and the number of building permits issued in each ward independently.
+Then we will fit a Bayesian regression model with varying intercepts to estimate the effect of these variables on the number of building permits issued in each ward.
 
-We employed multilevel modeling to estimate the effect of these variables on the number of building permits issued in each ward.
-Here we briefly describe a Bayesian Poisson regression model, which incorporates varying intercepts by grouping the effect by ward account.
-This approach allows us to capture the nuanced effects of demographic and economic factors on infrastructure development and resource distribution.
-Background details and diagnostics are included in [Appendix @sec-model-details].
+## Model set-up
 
-### Model set-up
-
-The model was specified as follows:
+The particular model we used is Bayesian multiple linear regression model with varying intercepts. The model includes the following variables:
 
 \begin{align*} 
-\text{Building Permits}_{i} & \sim \text{Poisson}(\lambda_{i}) \\
-\log(\lambda_{i}) & = \beta_{0} + \beta_{1} \times \text{Population}_{i} + \beta_{2} \times \text{Income}_{i} + \beta_{3} \times \text{Budget}_{i} + \alpha_{\text{ward}[i]} \\
-\beta_{0} & \sim \text{Normal}(0, 2.5) \\
-\beta_{1} & \sim \text{Normal}(0, 2.5) \\
-\beta_{2} & \sim \text{Normal}(0, 2.5) \\
-\beta_{3} & \sim \text{Normal}(0, 2.5) \\
-\alpha_{\text{ward}} & \sim \text{Normal}(0, \sigma_{\text{ward}})
+y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
+\mu_i &= \beta_0 + \beta_1 x_{\text{population}} + \beta_2 x_{\text{income}} + \beta_3 x_{\text{budget}} + \gamma_i \\
+\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
+\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
+\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
+\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
+\gamma_i &\sim \mbox{Normal}(0, 2.5) \\
+\sigma &\sim \mbox{Exponential}(1)
 \end{align*}
@@
 
--   $\text{Building Permits}_i$ represents the number of building permits issued in ward `i`,
--   $\lambda_i$ is the expected count of permits for ward $i$.
--   $\log(\lambda_i)$ is the linear predictor, which includes:
-    -   $\beta_0$: Intercept
-    -   $\beta_1$: Effect of `population`
-    -   $\beta_2$: Effect of `average_household_income`
-    -   $\beta_3$: Effect of `budget_allocation`
-    -   $\alpha_{\text{ward}[i]}$: Random effect for each ward (captures unobserved heterogeneity between wards).
--   Priors for $\beta_0$, $\beta_1$, $\beta_2$, and $\beta_3$ are Normal(0, 2.5), allowing for flexibility.
+- $\mu_i$ is the expected number of building permits issued in ward $i$.
+- $\beta_0$ is the intercept, representing the expected number of building permits issued in a ward with average values for all other variables.
+- $\beta_1$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the population density of the ward.
+- $\beta_2$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the average household income of the ward.
+- $\beta_3$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the total budget allocation across three categories: Growth Related, State of Good Repair, and Service Improvement and Enhancement.
+
+The model also includes a varying intercept $\gamma_i$ for each ward, accounting for the unobserved heterogeneity between wards.
+The varying intercepts allow the model to capture the unique characteristics of each ward that may influence the number of building permits issued.
 
@@
 The model also included a prior for the intercept, which was set to a normal distribution with a mean of 0 and a standard deviation of 2.5.
-The varying intercepts for each ward were modeled using a normal distribution with a mean of 0 and a standard deviation of $\sigma_{\text{ward}}$.
+The standard deviation of the likelihood was set to an exponential distribution with a rate parameter of 1.
 
-### Model justification
+## Model justification
+
+The above factors were choosen as they are known to influence the number of building permits issued in each ward.
+Population density and average household income are key demographic and economic indicators that can impact construction activity and development.
+Budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects are critical determinants of infrastructure development and resource distribution.
+By incorporating these variables into the model, we can assess their individual and combined effects on the number of building permits issued in each ward.
+
+A priori, we expect that population density and average household income will have a positive effect on the number of building permits issued, reflecting the demand for construction and development in densely populated and affluent areas.
+Similarly, we anticipate that budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects will positively influence the number of building permits, reflecting the city's investment in infrastructure development and maintenance.
 
@@
 
-## Results {#sec-results}
+# Results {#sec-results}
 
@@
 
-### Relationship between Population, Average Household Income, and Budget Allocations
-
-Our budget data spans three categories—`Growth Related`, `State of Good Repair`, and `Service Improvement and Enhancement`—for the years 2021-2024.
-As observed in [@sec-budget-trends], there has been a steady increase in budget allocations across these categories.
-To better understand the relationship between these budget categories and our variables of interest, we calculate the average budget allocation for each category across the years.
-The results are presented in [@tbl-avg-budget-data].
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-#| label: tbl-avg-budget-data
-#| tbl-cap: "Average 10-year budget allocations by category (2021-24)"
-#| tbl-pos: "h"
-
-# Calculate the average 10-year budget allocations by category
-average_budget <- analysis_budget_data |>
-  group_by(ward_id, ward, category) |>
-  summarise(avg_total_10_year = mean(total_10_year)) |>
-  pivot_wider(names_from = category, values_from = avg_total_10_year) |>
-  select(ward_id, ward, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`)
-
-average_budget |>
-  head() |>
-  kable(col.names = c("Ward ID", "Ward Name", "Growth Related", 
-                      "State of Good Repair", "Service Improvement and Enhancement"))
-```
-## Relationship between Average Household Income & Population with Budget Allocation for Growth Related Projects
+## Relationship between Building Permits and Population, Income, and Budget Allocations
 
@@
 #| warning: false
-#| label: fig-growth-related-budget
-#| fig.cap: "Growth Related Budget by Ward"
+#| label: fig-building-permits-relationship
+#| fig.cap: "Relationship between Variables and Total Building Permits"
+#| fig-subcap: ["Population", "Average Household Income", "Total Budget in Three Categories"]
 #| fig-pos: "h"
-#| fig.width: 10
-#| fig.height: 5
+#| layout-ncol: 2
 
-combined_data <- analysis_ward_data |>
-  left_join(average_budget, by = "ward_id") |>
+results_data <-
+  analysis_budget_data |>
+  group_by(ward_id, ward, category) |>
+  summarise(total_10_year = mean(total_10_year)) |>
+  pivot_wider(names_from = category, values_from = total_10_year) |>
+  select(ward_id, ward, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
+  left_join(analysis_building_permits, by = "ward_id") |>
+  left_join(analysis_ward_data, by = "ward_id") |>
   rename(ward = ward.x) |>
-  select(ward_id, ward, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`)
+  select(ward_id, ward, population, average_household_income, total_building_permits, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
+  # add a new column for total budget across three categories
+  mutate(total_budget = `Growth Related` + `State of Good Repair` + `Service Improvement and Enhancement`) |>
+  select(-`Growth Related`, -`State of Good Repair`, -`Service Improvement and Enhancement`)
 
-growth_budget <- combined_data |>
-  select(ward_id, ward, population, average_household_income, `Growth Related`)
+# Create multiple scatter plots
+plot1 <- ggplot(results_data, aes(x = population, y = total_building_permits, label = ward_id)) +
+    geom_point(color = "blue") +
+    geom_text_repel(hjust = -0.5, vjust = 0.5) +
+    geom_smooth(method = "lm", color = "red", se = TRUE) +
+    labs(x = "Population", y = "Total Building Permits")
 
-p1 <- ggplot(growth_budget, aes(x = population, y = `Growth Related`, label= ward_id)) +
-  geom_point() +
+plot2 <- ggplot(results_data, aes(x = average_household_income, y = total_building_permits, label = ward_id)) +
+    geom_point(color = "blue") +
     geom_text_repel(hjust = -0.5, vjust = 0.5) +
-  geom_smooth(method = "lm", color = "blue", se = TRUE) +
-  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
-  theme_minimal()
+    geom_smooth(method = "lm", color = "red", se = TRUE) +
+    labs(x = "Average Household Income", y = "Total Building Permits")
 
-p2 <- ggplot(growth_budget, aes(x = average_household_income, y = `Growth Related`, label= ward_id)) +
-  geom_point() +
+plot3 <-
+  ggplot(results_data, aes(x = total_budget, y = total_building_permits, label = ward_id)) +
+    geom_point(color = "blue") +
     geom_text_repel(hjust = -0.5, vjust = 0.5) +
-  geom_smooth(method = "lm", color = "blue", se = TRUE) +
-  labs(x = "Average Household Income", y = "") +
-  theme_minimal()
+    geom_smooth(method = "lm", color = "red", se = TRUE) +
+    labs(
+      x = "Total Budget in Three Categories (Sum)",
+      y = "Total Building Permits",
+    )
 
-# Combine both plots side by side
-grid.arrange(p1, p2, ncol = 2)
-
+plot1
+plot2
+plot3
 ```
 
-## Relationship between Average Household Income & Population with Budget Allocation for State of Good Repair Projects
+We now explore the results of independent comparisons between the outcome variable (total number of building permits in Toronto's 25 wards) and the predictor variables (population density, average household income, and budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects).
+As shown in [@fig-building-permits-relationship], most predictor variables exhibit relatively flat linear regression lines, indicating a weak relationship with the total number of building permits.
 
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-state-of-good-repair-budget
-#| fig.cap: "State of Good Repair Budget by Ward"
-#| fig-pos: "h"
-#| fig.width: 10
-#| fig.height: 5
+For example, the relationship between population density and building permits is notably flat, suggesting that population density has minimal influence on construction activity and development across the city’s wards—an unexpected result.
+A similar trend is observed for Growth-Related and State of Good Repair budget allocations, where flat regression lines suggest a weak relationship.
+This appears contradictory to earlier findings, where higher population density corresponded to increased budget allocations for these categories, which are typically associated with infrastructure development and maintenance.
+The same weak correlation is seen for Service Improvement and Enhancement budget allocations, as indicated by the flat regression line.
 
-good_repair_budget <- combined_data |>
-  select(ward_id, ward, population, average_household_income, `State of Good Repair`)
+In contrast, the relationship between average household income and building permits shows a strong positive correlation.
+Higher-income areas tend to experience more construction activity and development.
+For instance, Ward 15 (Don Valley West) has the highest number of building permits, aligning with its high average household income.
+This suggests that higher-income areas are more likely to see increased construction and development, highlighting a clear link between income levels and building permits.
 
-p1 <- ggplot(good_repair_budget, aes(x = population, y = `State of Good Repair`, label= ward_id)) +
-  geom_point() +
-  geom_text_repel(hjust = -0.5, vjust = 0.5) +
-  geom_smooth(method = "lm", color = "blue", se = TRUE) +
-  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
-  theme_minimal()
-
-p2 <- ggplot(good_repair_budget, aes(x = average_household_income, y = `State of Good Repair`, label= ward_id)) +
-  geom_point() +
-  geom_text_repel(hjust = -0.5, vjust = 0.5) +
-  geom_smooth(method = "lm", color = "blue", se = TRUE) +
-  labs(x = "Average Household Income", y = "") +
-  theme_minimal()
-
-# Combine both plots side by side
-grid.arrange(p1, p2, ncol = 2)
-```
-
-### Relationship between Average Household Income & Population with Budget Allocation for Service Improvement and Enhancement Projects
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-service-improvement-budget
-#| fig.cap: "Service Improvement and Enhancement Budget by Ward"
-#| fig-pos: "h"
-#| fig.width: 10
-#| fig.height: 5
-
-s_i_budget <- combined_data |>
-  select(ward_id, ward, population, average_household_income, `Service Improvement and Enhancement`)
-
-p1 <- ggplot(s_i_budget, aes(x = population, y = `Service Improvement and Enhancement`, label= ward_id)) +
-  geom_point() +
-  geom_text_repel(hjust = -0.5, vjust = 0.5) +
-  geom_smooth(method = "lm", color = "blue", se = TRUE) +
-  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
-  theme_minimal()
-
-p2 <- ggplot(s_i_budget, aes(x = average_household_income, y = `Service Improvement and Enhancement`, label= ward_id)) +
-  geom_point() +
-  geom_text_repel(hjust = -0.5, vjust = 0.5) +
-  geom_smooth(method = "lm", color = "blue", se = TRUE) +
-  labs(x = "Average Household Income", y = "") +
-  theme_minimal()
-
-# Combine both plots side by side
-grid.arrange(p1, p2, ncol = 2)
-```
+## Model Results {#sec-model-results}
 
 # Discussion {#sec-discussion}
+
+## Average Household Income
+
+# Weaknesses and Limitations {#sec-limitations}
+
+There are number of limitations that neeeds to be addressed in reagrds to the analysis.
+As we discussed in section @sec-data, the data used in this analysis is based on publicly available datasets from the City of Toronto.
+While these datasets provide valuable insights into the city's demographics, budget allocations, and building permits, they may not capture the full complexity of the city's infrastructure development and resource distribution.
+For example, the building permits dataset only includes active permits, which may not fully represent the scope of construction activity in Toronto.
+Additionally, the budget allocations dataset may not capture all infrastructure projects and expenditures, leading to potential gaps in the analysis.
+
+Additionally, as shown in @sec-appendix, that there are 45 unique statuses assigned to each submitted permit.
+For this analysis, we just focused on one subset of these statuses.
+This may not fully capture the complexity of the building permit process and the various stages of construction and development in Toronto.
+Future analyses could explore additional permit statuses to gain a more comprehensive understanding of construction activity in the city.
+
+One limitation of this study is the lack of detailed data on specific projects within each budget category, which could provide further insights into why certain wards receive more or less funding.
+Additionally, while this analysis focused on population density and average household income, other factors like infrastructure needs, community priorities, and political considerations likely play a role in budget decisions.
+Future research should explore these variables to build a more complete picture of how municipal budgets are allocated.
+
+A potential limitation is regard ot the model.
+We assumed that the relationship between population density, average household income, and budget allocations is linear, which may not fully capture the complexity of these relationships.
+We also assumed independence of each ward, which may not hold true in practice due to shared resources, infrastructure projects, and other factors that could influence building permit issuance.
+Future research could explore more sophisticated modeling techniques to account for these dependencies and non-linear relationships.
+
+# Future Work {#sec-future-work}
+
+This analysis provides a foundation for future research on infrastructure development, resource distribution, and economic well-being in Toronto.
+Future studies could explore additional factors that influence budget allocations and building permit issuance, such as infrastructure needs, community priorities, and political considerations.
+By incorporating these variables into the analysis, researchers can gain a more comprehensive understanding of how municipal budgets are allocated and how construction activity is distributed across the city's wards.
+
+In terms of policy recommendations, the city should adopt equity-based budgeting to ensure underfunded, high-density wards receive adequate resources, particularly for essential services like health and safety.
+Toronto could also implement a baseline funding model for high-need areas and enhance public engagement in the budget process, ensuring that residents of under served wards have a voice in resource allocation.
+These actions would promote more transparent, balanced, and equitable urban governance.
 
@@
 
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-# select only growth related, state of good repair and service improvement and enhancement
-service_improvement <- analysis_budget_data |>
-  filter(category == "Service Improvement and Enhancement") |>
-  mutate(total_10_year = total_10_year / 1000) |>
-  # remove negative values
-  filter(total_10_year > 0) |>
- rename(total_budget = total_10_year)
-
-# Calculate the outliers using IQR
-outliers <- service_improvement |>
-  group_by(year) |>
-  mutate(
-    Q1 = quantile(total_budget, 0.25),
-    Q3 = quantile(total_budget, 0.75),
-    IQR = Q3 - Q1,
-    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
-  ) |>
-  filter(is_outlier)
-
-# Plot with labels for outliers
-ggplot(service_improvement, aes(x = factor(year), y = total_budget)) +
-  geom_boxplot(fill = "skyblue", color = "darkblue") +
-  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
-  geom_text(
-    data = outliers, 
-    aes(label = ward), 
-    nudge_y = 15, 
-    check_overlap = TRUE
-  ) +  # Label only outliers
-  labs(
-    title = "Total Budget Over Years",
-    x = "Year",
-    y = "Total Budget (in thousands)"
-  ) +
-  theme_minimal() +
-  # remove legend
-  theme(legend.position = "none")
-```
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-# select only growth related, state of good repair and service improvement and enhancement
-growth_related <- analysis_budget_data |>
-  filter(category == "Growth Related") |>
-  mutate(total_10_year = total_10_year / 1000) |>
-  # remove negative values
-  filter(total_10_year > 0) |>
- rename(total_budget = total_10_year)
-
-# Calculate the outliers using IQR
-outliers <- growth_related |>
-  group_by(year) |>
-  mutate(
-    Q1 = quantile(total_budget, 0.25),
-    Q3 = quantile(total_budget, 0.75),
-    IQR = Q3 - Q1,
-    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
-  ) |>
-  filter(is_outlier)
-
-# Plot with labels for outliers
-ggplot(growth_related, aes(x = factor(year), y = total_budget)) +
-  geom_boxplot(fill = "skyblue", color = "darkblue") +
-  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
-  geom_text(
-    data = outliers, 
-    aes(label = ward), 
-    nudge_y = 15, 
-    check_overlap = TRUE
-  ) +  # Label only outliers
-  labs(
-    title = "Total Budget Over Years",
-    x = "Year",
-    y = "Total Budget (in thousands)"
-  ) +
-  theme_minimal() +
-  # remove legend
-  theme(legend.position = "none")
-```
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-# select only growth related, state of good repair and service improvement and enhancement
-state_of_good_repair <- analysis_budget_data |>
-  filter(category == "State of Good Repair") |>
-  mutate(total_10_year = total_10_year / 1000) |>
-  # remove negative values
-  filter(total_10_year > 0) |>
- rename(total_budget = total_10_year)
-
-# Calculate the outliers using IQR
-outliers <- state_of_good_repair |>
-  group_by(year) |>
-  mutate(
-    Q1 = quantile(total_budget, 0.25),
-    Q3 = quantile(total_budget, 0.75),
-    IQR = Q3 - Q1,
-    is_outlier = total_budget < (Q1 - 1.5 * IQR) | total_budget > (Q3 + 1.5 * IQR)
-  ) |>
-  filter(is_outlier)
-
-# Plot with labels for outliers
-ggplot(state_of_good_repair, aes(x = factor(year), y = total_budget)) +
-  geom_boxplot(fill = "skyblue", color = "darkblue") +
-  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
-  geom_text(
-    data = outliers, 
-    aes(label = ward), 
-    nudge_y = 15, 
-    check_overlap = TRUE
-  ) +  # Label only outliers
-  labs(
-    title = "Total Budget Over Years",
-    x = "Year",
-    y = "Total Budget (in thousands)"
-  ) +
-  theme_minimal() +
-  # remove legend
-  theme(legend.position = "none")
-```
-
 # References

### 2024-12-03 scripts/00-simulate_data.R

@@ -1,52 +1,56 @@
 #### Preamble ####
-# Purpose: Simulates a dataset of Australian electoral divisions, including the 
-  #state and party that won each division.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Simulate data for the project
+# Author: Aviral Bhardwaj
+# Date: 2024-12-01
+# Contact: aviral.bhardwaj@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The `tidyverse` package must be installed
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-
+# Pre-requisites: N.A
+# Any other information needed? N.A
 
 #### Workspace setup ####
 library(tidyverse)
-set.seed(853)
+set.seed(9407)
 
 
 #### Simulate data ####
-# State names
-states <- c(
-  "New South Wales",
-  "Victoria",
-  "Queensland",
-  "South Australia",
-  "Western Australia",
-  "Tasmania",
-  "Northern Territory",
-  "Australian Capital Territory"
-)
+ward_id <- tibble(ward = 1:25)
+ward_names <- c("Etobicoke North", "Etobicoke Centre", "Etobicoke-Lakeshore",
+                "Parkdale-High Park","York South-Weston", "York Centre",
+                "Humber River-Black Creek", "Eglinton-Lawrence",
+                "Davenport", "Spadina-Fort York", "University-Rosedale",
+                "Toronto-St. Paul's", "Toronto Centre", "Toronto-Danforth",
+                "Don Valley West", "Don Valley East", "Don Valley North",
+                "Willowdale", "Beaches-East York", "Scarborough Southwest",
+                "Scarborough Centre", "Scarborough-Agincourt",
+                "Scarborough North", "Scarborough-Guildwood",
+                "Scarborough-Rouge Park")
 
-# Political parties
-parties <- c("Labor", "Liberal", "Greens", "National", "Other")
+ward_id <- ward_id |>
+  mutate(ward_name = ward_names) |>
+  rename(ward_id = ward) |>
+  rename(ward = ward_name)
 
-# Create a dataset by randomly assigning states and parties to divisions
-analysis_data <- tibble(
-  division = paste("Division", 1:151),  # Add "Division" to make it a character
-  state = sample(
-    states,
-    size = 151,
-    replace = TRUE,
[…]

### 2024-12-03 scripts/01-test_simulated_data.R

@@ -1,89 +1,153 @@
 #### Preamble ####
-# Purpose: Tests the structure and validity of the simulated Australian 
-  #electoral divisions dataset.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Test the simulated data for the project
+# Author: Aviral Bhardwaj
+# Date: 2024-12-01
+# Contact: aviral.bhardwaj@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
   # - The `tidyverse` package must be installed and loaded
   # - 00-simulate_data.R must have been run
-# Any other information needed? Make sure you are in the `starter_folder` rproj
+# Any other information needed? Make sure you are in the correct working directory.
 
 
 #### Workspace setup ####
 library(tidyverse)
 
-analysis_data <- read_csv("data/00-simulated_data/simulated_data.csv")
+# Load the simulated data
+ward_data <- read_csv(here::here("data/00-simulated_data/ward_data.csv"))
+analysis_budget <- read_csv(here::here("data/00-simulated_data/analysis_budget.csv"))
+building_permits <- read_csv(here::here("data/00-simulated_data/building_permits.csv"))
 
 # Test if the data was successfully loaded
-if (exists("analysis_data")) {
-  message("Test Passed: The dataset was successfully loaded.")
+if (all(c("ward_data", "analysis_budget", "building_permits") %in% ls())) {
+  message("Test Passed: Data loaded successfully.")
 } else {
-  stop("Test Failed: The dataset could not be loaded.")
+  stop("Test Failed: Data not loaded.")
 }
 
 
 #### Test data ####
-
-# Check if the dataset has 151 rows
-if (nrow(analysis_data) == 151) {
-  message("Test Passed: The dataset has 151 rows.")
+# 1. Row Count Check for `ward_id`
+if (nrow(ward_id) == 25) {
+  print("Test Passed: `ward_id` has exactly 25 rows")
 } else {
-  stop("Test Failed: The dataset does not have 151 rows.")
+  print("Test Failed: `ward_id` does not have 25 rows")
 }
 
-# Check if the dataset has 3 columns
-if (ncol(analysis_data) == 3) {
-  message("Test Passed: The dataset has 3 columns.")
+# 2. Column Count Check for `ward_id`
+if (ncol(ward_id) == 2) {
+  print("Test Passed: `ward_id` has exactly 2 columns")
 } else {
-  stop("Test Failed: The dataset does not have 3 columns.")
[…]

### 2024-12-03 scripts/03-clean_data.R

@@ -183,7 +183,7 @@ building_permits <- building_permits |>
 Application Accepted" | STATUS == "Issuance Pending" | STATUS == "Ready for Issuance" | STATUS == "Permit Issued" | STATUS == "Permit Issued/Close File") |>
   filter(!is.na(POSTAL)) |>
   filter(!is.na(STATUS)) |>
-  filter(grepl("new building", WORK, ignore.case = TRUE)) |>
+  # filter(grepl("new building", WORK, ignore.case = TRUE)) |>
   left_join(postal_code_to_ward, by = c("POSTAL" = "postal_code")) |>
   filter(!is.na(primary_ward)) |>
   rename("ward_id" = "primary_ward")

### 2024-12-03 scripts/04-test_analysis_data.R

@@ -1,69 +1,141 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 26 September 2024 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Test the simulated data for the project
+# Author: Aviral Bhardwaj
+# Date: 2024-12-01
+# Contact: aviral.bhardwaj@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 
+# - The `tidyverse` package must be installed and loaded
+# - 02-download_data.R must be run before this script
+# - 03-clean_data.R must be run before this script
+# Any other information needed? Make sure you are in the correct working directory.
 
 
 #### Workspace setup ####
 library(tidyverse)
 library(testthat)
+library(readr)
+library(dplyr)
 
-data <- read_csv("data/02-analysis_data/analysis_data.csv")
-
+# Load the cleaned analysis data
+ward_data <-read.csv(here::here("data/02-analysis_data/analysis_ward_data.csv"))
+budget_data <- read.csv(here::here("data/02-analysis_data/analysis_budget_data.csv"))
+building_permits <- read.csv(here::here("data/02-analysis_data/analysis_building_permits.csv"))
 
 #### Test data ####
-# Test that the dataset has 151 rows - there are 151 divisions in Australia
-test_that("dataset has 151 rows", {
-  expect_equal(nrow(analysis_data), 151)
+test_that("Budget Data Integrity Tests", {
+  expect_true(all(is.integer(budget_data$ward_id)))
+  expect_true(all(budget_data$year %in% c(2021, 2022, 2023, 2024)))
+  expect_false(any(is.na(budget_data$ward_id)))
+  expect_false(any(is.na(budget_data$ward)))
+  expect_false(any(is.na(budget_data$category)))
+  expect_false(any(is.na(budget_data$total_10_year)))
+  expect_false(any(is.na(budget_data$year)))
 })
 
-# Test that the dataset has 3 columns
-test_that("dataset has 3 columns", {
-  expect_equal(ncol(analysis_data), 3)
+test_that("Building Permits Data Tests", {
+  expect_equal(ncol(building_permits), 2)
[…]

### 2024-12-04 paper/dump.qmd
Complete Paper commit

@@
 
-model <- stan_glm(
-  formula = total_building_permits ~ total_budget + average_household_income + population + ward_id,
+model <- stan_glmer(
+  formula = total_building_permits ~ total_budget + average_household_income + (1 | ward_id),
   data = model_data,
@@
 
+# posterior predictive vs observed
+posterior_vs_prior(model) +
+  theme_minimal() +
+  scale_color_brewer() +
+  theme(legend.position = "bottom") +
+  coord_flip()
+
 # plot diagnostics

### 2024-12-04 paper/paper.qmd
Complete Paper commit

@@
 ---
-title: "Who Gets What: How Population and Income Shape Toronto’s Budget Decisions"
-subtitle: "Understanding the Impact of Demographics and Economic Factors on City Spending Across 25 Wards"
+title: "Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto"
+subtitle: "Population Density Drive Larger Budget Allocations meanwhile Average Household Income Drives Construction Projects"
 author: 
@@
 date-format: long
-abstract: "This study investigates how population density and average household income affect budget allocations across Toronto’s 25 wards.
-Using data from the city's 2023-2032 Capital Budget Plan and 2021 Ward Profiles, we analyze spending patterns in key areas like health and safety, infrastructure, and community services.
-Our findings reveal that higher population densities do not guarantee proportionally greater investments in essential services, while wealthier wards often receive larger capital expenditures.
+abstract: "This study investigates how population density and average household income affect budget allocations across Toronto’s 25 wards, and how these factors, along with budget allocation, influence construction projects.
+Using data from the city's \"Capital Budget Plan\", \"Ward Profiles\" and \"Active Building Permits\", we analyze spending patterns in key areas like related to infrastructre.
+Our analysis, guided by a causal model, reveals a negative association between population density and construction activity, with some high-density wards receiving significantly higher budget allocations.
+Average household income and total budget allocations were positively correlated with the number of building permits.
 These insights highlight potential disparities in resource distribution, emphasizing the need for more equitable urban budget planning to address diverse community needs."
@@
 library(gridExtra)
+library(modelsummary)
+library(rstanarm)
+library(bayesplot)
 ```
@@
 # Get Model
-# model <- readRDS(here::here("models/model.rds"))
+model <- readRDS(here::here("models/model.rds"))
 ```
@@
 # Introduction {#sec-intro}
+
+In urban governance, the equitable allocation of municipal budgets is essential to meet a city’s social, economic, and infrastructure needs.
+This challenge is especially pressing in Toronto, where 25 wards compete for limited resources amid a growing population.
+Understanding how demographic factors like population and income influence budget decisions remains underexplored.
+This study seeks to answer: How do population and income affect budget allocations across Toronto’s wards, and do these factors, along with budget allocation, impact the number of construction projects, a key indicator of infrastructure development?
+
+Historical budget data shows uneven investment, particularly in high-density, lower-income areas, leading to disparities in infrastructure and services [@ferguson2024compromised].
+Following her 2023 election, Mayor Olivia Chow prioritized housing, transit, and equity-focused initiatives [@toronto_2024_budget].
+The 2024 budget reflects these priorities, with significant investments in affordable housing, transportation, and community safety, while emphasizing long-term infrastructure development through a 10-year capital plan.
+However, it remains unclear if these allocations are equitably distributed across wards or align with Toronto’s broader economic and demographic realities.
+
+Major projects like the Ontario Line—a new subway connecting Toronto’s east and west ends—are expected to spur economic growth and development [@citeTransitExpansion].
+Similarly, Olivia Chow’s "Sidewalks to Skylines" 10-year plan aims to create jobs and improve residents' quality of life [@citeSidewalksToSkylines].
+As these initiatives unfold, equitable infrastructure investments will be vital for Toronto’s long-term prosperity.
+With more residents expected to spread across the city's wards, it is essential to evaluate whether current budget allocations align with the city’s broader development goals as the demographic characteristics of each ward changes.
+
+This study uses data from the City of Toronto’s Open Data Portal, including demographic information from the "2021 Ward Profiles", "Capital Budget & Plan By Ward (10 yr Approved) 2021-2024", and "Building Applications and Permits".
+Our analysis, guided by a causal model, shows that both population density and average household income significantly influence budget allocations and the issuance of building permits, key indicators of infrastructure development.
+However, population density was identified as an exclusion restriction, as its inclusion in the Bayesian multiregression model revealed a counterintuitive negative relationship with the number of building permits.
+This suggests that unobserved factors may be influencing the link between population density and construction activity.
+After excluding population density from the model, we found that average household income and total budget allocations were positively correlated with building permits.
+This result aligns with expectations: wealthier wards with higher incomes receive larger budget allocations, leading to more construction projects.
+
+The paper is organized as follows: [@sec-data] outlines the data sources and methodology used in the analysis.
+[@sec-model] explains the Bayesian model applied to estimate the effects of demographic and economic factors, along with budget allocations, on construction projects.
+[@sec-results] presents key findings, including trends in budget allocations, the relationship between demographic factors and budget distribution, and the impact on construction projects.
+[@sec-discussion] explores the implications for urban governance and budget planning in Toronto.
+Finally, [@sec-limitations] addresses the study’s limitations and [@sec-future-work] offers recommendations for future research.
+Additional details, including tables, figures, and methodology, are provided in the [@sec-appendix].
+
+## Estimand
+
+Our research question and estimand analyse the causal relationships between population, income, budget allocation, and active construction projects ([@fig-causal-model]).
+Population and income serve as confounders, influencing both budget allocation and construction projects.
+Budget allocation acts as a mediator, connecting the confounders to the outcome, and as an instrumental variable, helping to estimate the effect of these predictors on the number of construction projects.
+The outcome variable is active construction projects, while population and income are the predictor variables.
+We aim to estimate the causal effect (estimand) of budget allocation on construction projects, while accounting for the influence of population and income.
+This paper highlights how budget decisions mediate the broader impact of socioeconomic factors on infrastructure development.
+
+Population is a critical demographic indicator, representing the number of residents in each ward, while average household income reflects economic well-being, influencing access to resources and quality of life [@Schaeffer].
+Therefore, these two variables of interest were chosen for the analysis.
+For budget allocations we choose these specific categories of interest, specifically Growth-Related expenditures, State of Good Repair, and Service Improvement and Enhancement.
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-causal-model
+#| fig-cap: "Causal model"
+#| fig-subcap: "Population and income influence budget allocation and construction projects"
+#| fig-pos: "h"
+
+# create a dag for the model
+causal_model <- "digraph {
+  graph []
+  # Nodes
+  node [shape = plaintext]
+    A [label = 'Population']
+    B [label = 'Budget Allocation']
+    C [label = 'Active Construction Projects']
+    D [label = 'Income']
+{ rank = same A D; }
+  # Edges for confounding variable
+  edge []
+    A -> {B C}
+    D -> {B C}
+    B -> C
+}"
+
+# render the causal model
+exporting_image <- grViz(causal_model) |>
+  export_svg() |>
+  charToRaw() |>
+  rsvg_png(here::here("other/sketches/causal_model.png"), width = 800, height = 600)
+
+image <- readPNG(here::here("other/sketches/causal_model.png"))
+par(mar = rep(2, 4))
+plot.new()
+rasterImage(image, 0, 0, 1, 1)  
+```
 
@@
 The `patchwork` [@patchwork] package was used for combining multiple plots, and `sf` [@sf] for spatial data analysis.
-For models, we used `rstan` [@rstan] was used for fitting the model, and `modelsummary` [@modelsummary] for model summary tables.
-The `lintr` [@lintr] package was used for code linting.
-The `arrow` [@arrow] package was used for reading and writing Parquet files.
-
-## Measurement
-
-Our research question and estimand analyse the relationship between key demographic and economic factors—specifically, population and average household income—and the budget allocation across various categories in Toronto's 25 wards.
-Population is a critical demographic indicator, representing the number of residents in each ward, while average household income reflects economic well-being, influencing access to resources and quality of life [@Schaeffer].
-Therefore, these two variables of interest were chosen for the analysis.
-Our primary aim is to estimate the impact of these variables on budget allocations for these specific categories of interest, specifically Growth-Related expenditures, State of Good Repair, and Service Improvement and Enhancement.
-
-Population density is expected to affect the demand for services and infrastructure, while average household income may shape how resources are distributed across wards, reflecting broader economic disparities.
-The estimand is the causal effect of these factors on the allocation of funds, and the estimator will quantify how shifts in population and income levels influence budgetary decisions.
-By understanding this relationship, we aim to offer insights into how demographic and economic variables drive municipal spending patterns in different categories across the city.
-In order to better analyze the trends and modelling the data, we will use the Captial Budget and Plan Details from 2021-2024 datasets, comparing total 10-year budget allocations across different wards and categories of interest.
+The `lintr` [@citeLintr] package was used for code linting.
 
@@
 
-Each budget plan includes five primary categories under `State of Good Repair`, `Growth Related`, `Health and Safety`, `Service Improvement and Enhancement`, and `Legislated`.
+Each budget plan includes five primary categories under *State of Good Repair*, *Growth Related*, *Health and Safety*, *Service Improvement and Enhancement*, and *Legislated*.
 These categories define the main areas where capital expenditures are directed.
-For this analysis, however, we will focus on these three variables of interest: `State of Good Repair`, `Growth Related`, and `Service Improvement and Enhancement`.
+For this analysis, however, we will focus on these three variables of interest: *State of Good Repair*, *Growth Related*, and *Service Improvement and Enhancement*.
 
 These categories were selected because they represent critical areas of investment that directly impact the quality of life and well-being of Toronto residents.
-`State of Good Repair` focuses on maintaining and preserving existing infrastructure, `Growth Related` addresses the expansion and development of new infrastructure, and `Service Improvement and Enhancement` aims to enhance public services and amenities.
+*State of Good Repair* focuses on maintaining and preserving existing infrastructure, *Growth Related* addresses the expansion and development of new infrastructure, and *Service Improvement and Enhancement* aims to enhance public services and amenities.
 By analyzing budget allocations in these categories, we can gain valuable insights into the city's spending priorities and identify opportunities for more effective and equitable resource distribution.
@@
 
-Table [@tbl-raw-building-permits-data] outlines detailed information about active building permits in Toronto, including key features such as `Permit Number`, `Permit Type`, `Structure Type`, `Status`, and more.
+[@tbl-raw-building-permits-data] outlines detailed information about active building permits in Toronto, including key features such as `Permit Number`, `Permit Type`, `Structure Type`, `Status`, and more.
 
@@
 
-To ensure we analyzed only construction-related permits, we filtered out entries where `Work` did not have the value `New Building`.
+<!-- To ensure we analyzed only construction-related permits, we filtered out entries where `Work` did not have the value `New Building`. -->
 The cleaned data was then saved in CSV and Parquet formats with the following columns:
@@
 
-## Relationship between Population, Average Household Income, and Budget Allocations
+The box plots in [@fig-budget-trends] illustrate the trends in budget allocations across three categories—*Growth Related*, *State of Good Repair*, and *Service Improvement and Enhancement*—from 2021 to 2024.
+Each box plot represents the distribution of budget allocations for each year, with the median value indicated by the horizontal line inside the box.
+Outliers are labeled with the corresponding ward ID, highlighting wards with unusually high or low budget allocations.
 
-Our budget data spans three categories—`Growth Related`, `State of Good Repair`, and `Service Improvement and Enhancement`—for the years 2021-2024.
+Ward 10 (Spadina-Fort York) received the highest budget allocations across both *Growth Related* and *State of Good Repair* categories, indicating significant investment in infrastructure development and maintenance to population-dense areas.
+Toronto Center, Don Valley East, and Etobicoke Center also received notable budget allocations, some of these wards have higher average household incomes as well as higher population densities.
+The box plots reveal that budget allocations for *Growth Related* projects have increased steadily over the years, with 75% of wards receiving higher allocations in 2024 compared to 2021.
+In contrast, budget allocations for *State of Good Repair* and *Service Improvement and Enhancement* projects have remained relatively consistent, with few outliers receiving higher budget allocations.
+
+The box plots for *State of Good Repair* and *Service Improvement and Enhancement* projects however, are less varied as most of the wards received similar budget allocations across the years, with few outliers such as University-Rosedale, Toronto Centre and Don Valley West receiving higher budget allocations, who coincidentally have higher average household income.
+
+## Relationship between Variables of Interest and Budget Allocations {#sec-rel-variables}
+
+Our budget data spans three categories—*Growth Related*, *State of Good Repair*, and *Service Improvement and Enhancement*—for the years 2021-2024.
 As observed in [@sec-budget-trends], there has been a steady increase in budget allocations across these categories.
@@
 ```
-### Relationship between Average Household Income & Population with Budget Allocation for Growth Related Projects and State of Good Repair Projects
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-growth-related-budget
-#| fig.cap: "Growth Related Budget by Ward"
-#| fig-pos: "h"
-#| fig.width: 10
-#| fig.height: 5
-
-combined_data <- analysis_ward_data |>
-  left_join(average_budget, by = "ward_id") |>
-  left_join(analysis_building_permits, by = "ward_id") |>
-  rename(ward = ward.x) |>
-  select(ward_id, ward, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`, total_building_permits)
-
-growth_budget <- combined_data |>
-  select(ward_id, ward, population, average_household_income, `Growth Related`)
-
-p1 <- ggplot(growth_budget, aes(x = population, y = `Growth Related`, label= ward_id)) +
-  geom_point() +
-  geom_text_repel(hjust = -0.5, vjust = 0.5) +
-  geom_smooth(method = "lm", color = "blue", se = TRUE) +
-  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
-  theme_minimal()
-
-p2 <- ggplot(growth_budget, aes(x = average_household_income, y = `Growth Related`, label= ward_id)) +
-  geom_point() +
-  geom_text_repel(hjust = -0.5, vjust = 0.5) +
-  geom_smooth(method = "lm", color = "blue", se = TRUE) +
-  labs(x = "Average Household Income", y = "") +
-  theme_minimal()
-
-# Combine both plots side by side
-grid.arrange(p1, p2, ncol = 2)
-
-```
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-state-of-good-repair-budget
-#| fig.cap: "State of Good Repair Budget by Ward"
-#| fig-pos: "h"
-#| fig.width: 10
-#| fig.height: 5
-
-good_repair_budget <- combined_data |>
-  select(ward_id, ward, population, average_household_income, `State of Good Repair`)
-
-p1 <- ggplot(good_repair_budget, aes(x = population, y = `State of Good Repair`, label= ward_id)) +
-  geom_point() +
-  geom_text_repel(hjust = -0.5, vjust = 0.5) +
-  geom_smooth(method = "lm", color = "blue", se = TRUE) +
-  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
-  theme_minimal()
-
-p2 <- ggplot(good_repair_budget, aes(x = average_household_income, y = `State of Good Repair`, label= ward_id)) +
-  geom_point() +
-  geom_text_repel(hjust = -0.5, vjust = 0.5) +
-  geom_smooth(method = "lm", color = "blue", se = TRUE) +
-  labs(x = "Average Household Income", y = "") +
-  theme_minimal()
-
-# Combine both plots side by side
-grid.arrange(p1, p2, ncol = 2)
-```
-[@fig-growth-related-budget] illustrates the relationship between average household income, population, and the total 10-year budget allocation for Growth-Related projects.
-The linear regression line for average household income remains relatively flat, hovering around \$100,000.
-This suggests that budget allocations for Growth-Related projects are not significantly influenced by income levels, indicating a more uniform distribution of funds across wards regardless of household income.
-A similar lack of correlation is observed in [@fig-state-of-good-repair-budget], where budget allocations for State of Good Repair projects also show minimal dependence on average household income.
-
-In contrast, the linear regression line for population shows a positive correlation with budget allocation for Growth-Related projects.
-Wards with higher population densities—such as Ward 10 (Spadina-Fort York), Ward 13 (Toronto Centre), Ward 5 (York South-Weston), and Ward 2 (Etobicoke North)—receive a larger share of the Growth-Related budget.
-This indicates that population density plays a critical role in determining funding levels for these projects, with more populous wards benefiting from increased allocations.
-A similar trend is observed in the budget distribution for State of Good Repair projects, where wards with higher population densities receive a proportionally larger share of the funds.
-
 ### Relationship between Average Household Income & Population with Budget Allocation for Service Improvement and Enhancement Projects 
@@
 #| fig.height: 5
+
+combined_data <- analysis_ward_data |>
+  left_join(average_budget, by = "ward_id") |>
+  left_join(analysis_building_permits, by = "ward_id") |>
+  rename(ward = ward.x) |>
+  select(ward_id, ward, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`, total_building_permits)
 
@@
 
+### Relationship between Average Household Income & Population with Budget Allocation for Growth Related Projects and State of Good Repair Projects
+
+[@fig-growth-related-budget] (a) illustrates the relationship between average household income, population, and the total 10-year budget allocation for Growth-Related projects.
+The linear regression line for average household income remains relatively flat, hovering around \$100,000.
+This suggests that budget allocations for Growth-Related projects are not significantly influenced by income levels, indicating a more uniform distribution of funds across wards regardless of household income.
+A similar lack of correlation is observed in [@fig-growth-related-budget] (b), where budget allocations for State of Good Repair projects also show minimal dependence on average household income.
+
+In contrast, the linear regression line for population shows a strong positive correlation with budget allocation for Growth-Related projects.
+Wards with higher population densities—such as Ward 10 (Spadina-Fort York), Ward 13 (Toronto Centre), Ward 5 (York South-Weston), and Ward 2 (Etobicoke North)—receive a larger share of the Growth-Related budget.
+This indicates that population density plays a critical role in determining funding levels for these projects, with more populous wards benefiting from increased allocations.
+A similar trend is observed in the budget distribution for State of Good Repair projects, where wards with higher population densities receive a proportionally larger share of the funds.
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-growth-related-budget
+#| fig.cap: "Growth Related Budget and State of Good Repair Budget by Ward"
+#| fig-subcap: ["Growth Related Budget", "State of Good Repair Budget"]
+#| fig-pos: "h"
+#| fig.width: 10
+#| fig.height: 5
+#| layout-ncol: 1
+
+growth_budget <- combined_data |>
+  select(ward_id, ward, population, average_household_income, `Growth Related`)
+
+p1 <- ggplot(growth_budget, aes(x = population, y = `Growth Related`, label= ward_id)) +
+  geom_point() +
+  geom_text_repel(hjust = -0.5, vjust = 0.5) +
+  geom_smooth(method = "lm", color = "blue", se = TRUE) +
+  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
+  theme_minimal()
+
+p2 <- ggplot(growth_budget, aes(x = average_household_income, y = `Growth Related`, label= ward_id)) +
+  geom_point() +
+  geom_text_repel(hjust = -0.5, vjust = 0.5) +
+  geom_smooth(method = "lm", color = "blue", se = TRUE) +
+  labs(x = "Average Household Income", y = "") +
+  theme_minimal()
+
+# Combine both plots side by side
+grid.arrange(p1, p2, ncol = 2)
+
+good_repair_budget <- combined_data |>
+  select(ward_id, ward, population, average_household_income, `State of Good Repair`)
+
+p1 <- ggplot(good_repair_budget, aes(x = population, y = `State of Good Repair`, label= ward_id)) +
+  geom_point() +
+  geom_text_repel(hjust = -0.5, vjust = 0.5) +
+  geom_smooth(method = "lm", color = "blue", se = TRUE) +
+  labs(x = "Population", y = "Average Total 10-Year Budget (in 000s) 2021-24") +
+  theme_minimal()
+
+p2 <- ggplot(good_repair_budget, aes(x = average_household_income, y = `State of Good Repair`, label= ward_id)) +
+  geom_point() +
+  geom_text_repel(hjust = -0.5, vjust = 0.5) +
+  geom_smooth(method = "lm", color = "blue", se = TRUE) +
+  labs(x = "Average Household Income", y = "") +
+  theme_minimal()
+
+# Combine both plots side by side
+grid.arrange(p1, p2, ncol = 2)
+
+```
+
+\newpage
+
 # Model {#sec-model}
@@
 y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \beta_0 + \beta_1 x_{\text{population}} + \beta_2 x_{\text{income}} + \beta_3 x_{\text{budget}} + \gamma_i \\
+\mu_i &= \beta_0 + \beta_1 x_{\text{income}} + \beta_2 x_{\text{budget}} + \gamma_i \\
 \beta_0 &\sim \mbox{Normal}(0, 2.5) \\
@@
 \beta_2 &\sim \mbox{Normal}(0, 2.5) \\
-\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
 \gamma_i &\sim \mbox{Normal}(0, 2.5) \\
@@
 - $\beta_0$ is the intercept, representing the expected number of building permits issued in a ward with average values for all other variables.
-- $\beta_1$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the population density of the ward.
-- $\beta_2$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the average household income of the ward.
-- $\beta_3$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the total budget allocation across three categories: Growth Related, State of Good Repair, and Service Improvement and Enhancement.
+- $\beta_1$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the average household income of the ward.
+- $\beta_2$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the total budget allocation across three categories: Growth Related, State of Good Repair, and Service Improvement and Enhancement.
 
@@
 
-The model was fitted using the `stan_glmer` function from the `rstanarm` package [@citerstanarm] in R [@citeR].
+The model was fitted using the `stan_glmer` function from the `rstanarm` package [@citerstanarm] in R [@citeR] and `modelsummary` [@citeModelsummary] for model summary tables.
 The prior distributions for the coefficients were set to normal distributions with a mean of 0 and a standard deviation of 2.5.
@@
 
-The above factors were choosen as they are known to influence the number of building permits issued in each ward.
-Population density and average household income are key demographic and economic indicators that can impact construction activity and development.
-Budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects are critical determinants of infrastructure development and resource distribution.
-By incorporating these variables into the model, we can assess their individual and combined effects on the number of building permits issued in each ward.
-
-A priori, we expect that population density and average household income will have a positive effect on the number of building permits issued, reflecting the demand for construction and development in densely populated and affluent areas.
-Similarly, we anticipate that budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects will positively influence the number of building permits, reflecting the city's investment in infrastructure development and maintenance.
-
-```{r}
-#| warning: false
-#| message: false
-#| echo: false
-#| label: fig-causal-model
-#| fig-cap: "Causal model for the relationship between population, income, and budget allocations"
-#| fig-pos: "h"
-
-# create a dag for the model
-causal_model <- "digraph {
-  graph []
-  # Nodes
-  node [shape = plaintext]
-    A [label = 'Population']
-    B [label = 'Budget Allocation']
-    C [label = 'Active Construction Projects']
-    D [label = 'Income']
-{ rank = same A D; }
-  # Edges for confounding variable
-  edge []
-    A -> {B C}
-    D -> {B C}
-    B -> C
-}"
-
-# render the causal model
-exporting_image <- grViz(causal_model) |>
-  export_svg() |>
-  charToRaw() |>
-  rsvg_png(here::here("other/sketches/causal_model.png"), width = 800, height = 600)
-
-image <- readPNG(here::here("other/sketches/causal_model.png"))
-par(mar = rep(2, 4))
-plot.new()
-rasterImage(image, 0, 0, 1, 1)  
-```
-
-# Results {#sec-results}
-
-In this section, we visualize the relationships between population density, average household income, budget allocations with total number of building permits across Toronto's 25 wards.
-Additionally, we present our model results, highlighting how these variables influence the number of building permits issued in each ward.
-
-## Relationship between Building Permits and Population, Income, and Budget Allocations
+To better understand the underlying data, [@eda-model-justification] provides a summary of the mean, median, and standard deviation for each variable used in the model.
+The predictor variables have large values and different scales, which can cause numerical instability and convergence issues in the model.
+To address this, we will standardize the predictor variables by subtracting the mean and dividing by the standard deviation.
+This will scale the variables consistently and improve both the model's convergence and interpretability.
+Additionally, we did not use a Bayesian Poisson regression model or negative binomial because the number of building permits is a continuous variable, not count data.
+Therefore, we chose to use a Bayesian multiple linear regression model.
+This model allows us to estimate the effect of population density, average household income, and budget allocations on the number of building permits issued in each ward giving us a well-rounded understanding of the relationships between these variables.
 
@@
 #| warning: false
-#| label: fig-building-permits-relationship
-#| fig.cap: "Relationship between Variables and Total Building Permits"
-#| fig-subcap: ["Population", "Average Household Income", "Total Budget in Three Categories"]
-#| fig-pos: "h"
-#| layout-ncol: 2
+#| label: eda-model-justification
+#| tbl-cap: "Mean, Median, and Standard Deviation of Each Variable"
+#| tbl-pos: "h"
 
@@
   select(-`Growth Related`, -`State of Good Repair`, -`Service Improvement and Enhancement`)
+
+# mean of all population, average household income, total budget and total building permits
+mean_population <- mean(results_data$population)
+mean_income <- mean(results_data$average_household_income)
+mean_budget <- mean(results_data$total_budget)
+mean_building_permits <- mean(results_data$total_building_permits)
+
+# median of all population, average household income, total budget and total building permits
+median_population <- median(results_data$population)
+median_income <- median(results_data$average_household_income)
+median_budget <- median(results_data$total_budget)
+median_building_permits <- median(results_data$total_building_permits)
+
+# standard deviation of all population, average household income, total budget and total building permits
+sd_population <- sd(results_data$population)
+sd_income <- sd(results_data$average_household_income)
+sd_budget <- sd(results_data$total_budget)
+sd_building_permits <- sd(results_data$total_building_permits)
+
+# Create a data frame with the mean, median, and standard deviation of each variable
+model_justification <- data.frame(
+  Variable = c("Population", "Average Household Income", "Total Budget", "Total Building Permits"),
+  Mean = c(mean_population, mean_income, mean_budget, mean_building_permits),
+  Median = c(median_population, median_income, median_budget, median_building_permits),
+  SD = c(sd_population, sd_income, sd_budget, sd_building_permits)
+)
+
+# Kable table
+model_justification |>
+  kable(col.names = c("Variable", "Mean", "Median", "Standard Deviation"))
+
+```
+
+The factors chosen for the model are known to influence the number of building permits issued in each ward.
+Population density and average household income are demographic and economic indicators that can affect construction activity.
+Budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects are key drivers of infrastructure development and resource distribution.
+Including these variables allows for assessing their individual and combined effects on the number of building permits issued.
+
+When population density was included, for the causual model, it exhibited exclusion restriction qualities, as it showed an unexpected negative relationship with building permits, leading to its exclusion for better interpretability.
+The model was then refitted without population density, focusing on average household income and budget allocations as predictors.
+
+# Results {#sec-results}
+
+In this section, we visualize the relationships between population density, average household income, budget allocations with total number of building permits across Toronto's 25 wards.
+Additionally, we present our model results, highlighting how these variables influence the number of building permits issued in each ward.
+
+## Distribution of Average Household Income and Population Density {#sec-distribution}
+
+The City of Toronto is divided into 25 wards, and the 2021 census data highlights significant disparities in population and average household income among them.
+As shown in [@fig-toronto-map-population], Spadina-Fort York (Ward 10) and Etobicoke-Lakeshore (Ward 3) have the highest population densities, with 135,400 and 139,920 residents, respectively.
+In contrast, Don Valley West (Ward 15) has a lower population density of 101,025 but boasts the highest average household income at \$224,800.
+
+Population distribution reveals a concentration in the western part of the city and downtown areas.
+Notably, wards like Etobicoke Centre (Ward 2) and Etobicoke-Lakeshore (Ward 3), situated farther from downtown, have lower costs of living and correspondingly lower average household incomes, such as Etobicoke North (Ward 2) with \$95,200.
+
+An interesting pattern emerges around Don Valley West (Ward 15), which has the highest average household income.
+Wards such as University-Rosedale (Ward 11), Toronto-St.
+Paul's (Ward 12), and Eglinton-Lawrence (Ward 8) are clustered together, indicating a geographical correlation among high-income areas.
+Conversely, the wards with the highest population densities do not overlap with those that have the highest average household incomes, highlighting distinct socioeconomic patterns within the city.
+
+## Relationship between Predictor Variables and Building Permits
+
+We now explore the results of independent comparisons between the outcome variable (total number of building permits in Toronto's 25 wards) and the predictor variables (population density, average household income, and budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects).
+As shown in [@fig-building-permits-relationship], most predictor variables exhibit relatively flat linear regression lines, indicating a weak relationship with the total number of building permits.
+
+For example, the relationship between population density and building permits is notably flat, suggesting that population density has minimal influence on construction activity and development across the city’s wards—an unexpected result.
+A similar trend is observed for Growth-Related and State of Good Repair budget allocations, where flat regression lines suggest a weak relationship.
+This appears contradictory to earlier findings, where higher population density corresponded to increased budget allocations for these categories, which are typically associated with infrastructure development and maintenance.
+The same weak correlation is seen for Service Improvement and Enhancement budget allocations, as indicated by the flat regression line.
+
+In contrast, the relationship between average household income and building permits shows a strong positive correlation.
+Higher-income areas tend to experience more construction activity and development.
+For instance, Ward 15 (Don Valley West) has the highest number of building permits, aligning with its high average household income.
+This suggests that higher-income areas are more likely to see increased construction and development, highlighting a clear link between income levels and building permits.
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-building-permits-relationship
+#| fig.cap: "Relationship between Variables and Total Building Permits"
+#| fig-subcap: ["Population", "Average Household Income", "Total Budget in Three Categories"]
+#| fig-pos: "h"
+#| layout-ncol: 2
 
@@
 
-We now explore the results of independent comparisons between the outcome variable (total number of building permits in Toronto's 25 wards) and the predictor variables (population density, average household income, and budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects).
-As shown in [@fig-building-permits-relationship], most predictor variables exhibit relatively flat linear regression lines, indicating a weak relationship with the total number of building permits.
+## Model Results {#sec-model-results}
 
-For example, the relationship between population density and building permits is notably flat, suggesting that population density has minimal influence on construction activity and development across the city’s wards—an unexpected result.
-A similar trend is observed for Growth-Related and State of Good Repair budget allocations, where flat regression lines suggest a weak relationship.
-This appears contradictory to earlier findings, where higher population density corresponded to increased budget allocations for these categories, which are typically associated with infrastructure development and maintenance.
-The same weak correlation is seen for Service Improvement and Enhancement budget allocations, as indicated by the flat regression line.
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: tbl-model-results
+#| tbl-cap: "Model Results"
+#| tbl-pos: "h"
 
-In contrast, the relationship between average household income and building permits shows a strong positive correlation.
-Higher-income areas tend to experience more construction activity and development.
-For instance, Ward 15 (Don Valley West) has the highest number of building permits, aligning with its high average household income.
-This suggests that higher-income areas are more likely to see increased construction and development, highlighting a clear link between income levels and building permits.
+# model summary
+modelsummary(
+  list(
+    "Model" = model
+  )
+)
+```
 
-## Model Results {#sec-model-results}
+The results of the Bayesian regression model, summarized in [@tbl-model-results], provide important insights into the relationship between total budget allocations, average household income, and the number of building permits issued across wards.
+
+The model explains 59.1% of the variance in building permits (R² = 0.591), indicating a moderate level of explanatory power.
+The coefficients suggest that both total budget (0.530) and average household income (0.350) have a positive and significant impact on the number of building permits issued, implying that areas with higher budget allocations and income levels are more likely to see increased construction activity.
+
+The inclusion of ward-specific random effects (σ² = 0.141) highlights the importance of accounting for variation between wards, suggesting that there are unobserved factors specific to each ward influencing permit issuance.
+Although the adjusted R² (0.369) indicates that the model's explanatory power is somewhat limited after accounting for the complexity of the predictors, the model still provides valuable insights into the role of economic factors in shaping construction activity.
+With an RMSE of 0.58, the model's predictions are reasonably accurate, although there is still room for improvement.
+
+In constructing the model, we assumed linearity and additive effects, acknowledging that these simplifications may not fully capture the interactions among variables.
+The model's limitations, such as the exclusion of population density due to its unexpected negative relationship, reflect a trade-off between interpretability and capturing complex dynamics.
+Nevertheless, the findings align with expectations that wealthier areas and those with greater budget allocations are more likely to experience construction growth, reaffirming the importance of economic resources in urban development patterns.
 
@@
 
-## Average Household Income
+In this section, we discuss the implications of our findings on urban development, economic well-being, and resource distribution in Toronto.
+We explore the role of budget allocations, average household income, and population density in shaping construction activity and infrastructure development across the city's 25 wards.
+Our analysis highlights the complex interplay between economic factors, demographic trends, and policy decisions, shedding light on the challenges and opportunities for equitable urban growth in Toronto.
+
+## Building Permits and Economic Well-Being
+
+Construction plays a pivotal role in the economy, driving both job creation and economic growth [@citeConstruction].
+One of the key indicators of construction activity is the number of building permits issued within a ward.
+Increased construction correlates with job opportunities, infrastructure development, and overall community well-being.
+For example, investment in residential and non-residential construction rose from 24.9% in 2021 to 29.5% in 2023, with projections indicating a further 1.5% increase in construction employment by 2024 [@citeConstructionEmployment].
+
+As highlighted in [@sec-model-results] Bayesian regression model reveals that average household income has a significant positive effect on the issuance of building permits.
+This suggests that construction activity tends to concentrate in higher-income areas, where demand for development is stronger.
+This finding underscores the role of income levels in driving construction and infrastructure development, emphasizing the need for targeted investments in lower-income areas to stimulate economic growth and enhance community well-being.
+
+For potential property investors in Toronto, these insights offer a strategic advantage.
+Investing in areas with higher average household incomes may be more profitable due to the increased likelihood of ongoing construction activity, which can drive property value appreciation and strengthen the local economy.
+While high-population-density areas are typically attractive for investment, our findings suggest that this may not always be the case.
+
+## Investing in Higher-Income Areas vs. High-Density Areas
+
+The interplay between population density, average household income, and infrastructure investment in Toronto highlights a complex challenge in achieving equitable resource distribution.
+As seen in [@sec-rel-variables] high-income areas often attract more construction activity, driven by market forces and residents’ ability to finance new developments.
+Whereas, only specific high-density areas like Ward 10 (Spadina-Fort York) and Ward 13 (Toronto Centre) receive substantial investment despite high population density, suggesting efforts to address this imbalance.
+Meanwhile, as discussed in [@sec-distribution], low-density and lower-income wards risk being underserved.
+
+These findings underscore the importance of equity-based budgeting, where low-density and lower-income areas should receive targeted investments to meet critical infrastructure needs.
+This approach aligns with urban development strategies that prioritize inclusivity.
+Notable examples include transit expansions such as the Ontario Line and Crosstown extensions [@citeTransitExpansion].The Ontario Line is expected to connect high-density areas like Ward 10 (Spadina-Fort York) and Ward 13 (Toronto Centre) to the broader transit network, which could stimulate further development and infrastructure investment in these wards, fostering a more balanced urban landscape and distributing population growth more evenly across the city.
+
+For future policy, Toronto could enhance equity by expanding initiatives such as eliminating parking minimums.
+This policy reduces construction costs, making development in high-density areas more feasible.
+Such measures would promote a more balanced and sustainable urban environment, ensuring that population growth is accompanied by appropriate infrastructure investments.
+
+## Budget Allocations and Construction Activity
+
+Our analysis shows a complex relationship between budget allocations and construction activity in Toronto's 25 wards.
+As discussed in [@sec-budget-trends] there has been a steady increases in budget allocations for key initiatives such as Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects from 2021 to 2024, the distribution of building permits shows only a good correlation with these allocations [@sec-model-results].
+This finding suggests that budget allocations alone may not be the sole driver of construction activity, as other factors like market conditions, regulatory frameworks, and community priorities also play a significant role.
+
+This discrepancy reflects broader challenges highlighted in Toronto's recent 10-year economic plan [@citeEconomicPlan], which aims to tackle pressing issues such as housing, inequality, and congestion.
+The plan emphasizes the need for strategic investments that go beyond conventional infrastructure spending.
+For example, the city has committed $35 million to cultural initiatives over the next decade, signaling a shift toward holistic urban development that prioritizes cultural and social well-being alongside physical infrastructure improvements.
+This shift may explain the weak correlation between budget allocations and building permits, as construction activity is influenced by a broader range of factors beyond traditional infrastructure spending.
+
+Moreover, the Ford government’s fall economic statement indicates a shrinking provincial deficit, suggesting a cautious fiscal environment that may limit how budgetary increases translate into tangible construction outputs [@citeFallEconomicStatement].
+These dynamics suggest that other factors, such as regulatory frameworks, market conditions, or demographic shifts, may be playing a more significant role than budgetary allocations alone in shaping Toronto’s construction landscape.
 
@@
 
-There are number of limitations that neeeds to be addressed in reagrds to the analysis.
-As we discussed in section @sec-data, the data used in this analysis is based on publicly available datasets from the City of Toronto.
-While these datasets provide valuable insights into the city's demographics, budget allocations, and building permits, they may not capture the full complexity of the city's infrastructure development and resource distribution.
-For example, the building permits dataset only includes active permits, which may not fully represent the scope of construction activity in Toronto.
-Additionally, the budget allocations dataset may not capture all infrastructure projects and expenditures, leading to potential gaps in the analysis.
+There are several limitations to the analysis that need to be addressed.
+As discussed in [@sec-data], the datasets used in this study are publicly available from the City of Toronto.
+While they provide valuable insights into the city's demographics, budget allocations, and building permits, they may not fully capture the complexity of Toronto’s infrastructure development and resource distribution.
+For example, the building permits dataset includes only active permits, which may not reflect the entire scope of construction activity.
+Additionally, the budget allocations dataset may exclude certain infrastructure projects and expenditures, leading to potential gaps in the analysis.
 
-Additionally, as shown in @sec-appendix, that there are 45 unique statuses assigned to each submitted permit.
-For this analysis, we just focused on one subset of these statuses.
-This may not fully capture the complexity of the building permit process and the various stages of construction and development in Toronto.
-Future analyses could explore additional permit statuses to gain a more comprehensive understanding of construction activity in the city.
+Furthermore, as detailed in [@sec-appendix], there are 45 unique statuses assigned to each submitted permit.
+This analysis focused on only one subset of these statuses, which may not fully capture the complexity of the building permit process and the various stages of construction.
+Future research could include additional permit statuses for a more comprehensive understanding of construction activity.
+Another limitation is the lack of detailed data on specific projects within each budget category.
+This information could offer further insights into why certain wards receive more or less funding.
 
-One limitation of this study is the lack of detailed data on specific projects within each budget category, which could provide further insights into why certain wards receive more or less funding.
-Additionally, while this analysis focused on population density and average household income, other factors like infrastructure needs, community priorities, and political considerations likely play a role in budget decisions.
-Future research should explore these variables to build a more complete picture of how municipal budgets are allocated.
-
-A potential limitation is regard ot the model.
-We assumed that the relationship between population density, average household income, and budget allocations is linear, which may not fully capture the complexity of these relationships.
-We also assumed independence of each ward, which may not hold true in practice due to shared resources, infrastructure projects, and other factors that could influence building permit issuance.
-Future research could explore more sophisticated modeling techniques to account for these dependencies and non-linear relationships.
+Lastly, the model assumes that the relationships between population density, average household income, and budget allocations are linear, which may oversimplify these relationships.
+Additionally, the analysis assumes independence among wards, which may not be accurate due to shared resources and infrastructure projects.
+Future research could explore more advanced modeling techniques to address these non-linear relationships and dependencies.
 
@@
 
-## Data Tables
-
-### Building Permits Status
+## Building Permits Status
 
@@
 
+## Posterior Predictive Check
+
+We employ a posterior predictive check in @fig-ppcheck to evaluate how well the model fits the data.
+By examining the distribution of the observed data against the simulated data, we can assess the model's predictive accuracy and identify any discrepancies.
+[@citeAlexander].
+
+```{r}
+#| label: fig-ppcheck
+#| fig-cap: Examining how the model fits the data through a posterior predictive check
+#| echo: false
+#| fig-pos: H
+#| message: false
+#| warning: false
+
+pp_check(model) + 
+  theme_classic() +
+  theme(legend.position = "bottom")
+```
+
+## Comparison of the Posterior and Prior
+
+We compare the posterior and prior in @fig-postprior to examine how the estimates change once data is taken into account [@citeAlexander].
+
+```{r}
+#| label: fig-postprior
+#| fig-cap: Examining how the model is affected by the data through a posterior-prior comparison
+#| echo: false
+#| warning: false
+#| fig-pos: H
+#| message: false
+#| fig-width: 10
+#| fig-height: 10
+
+model_data <- analysis_budget_data |>
+  group_by(ward_id, category, ward) |>
+  summarise(avg_total_10_year = mean(total_10_year)) |>
+  pivot_wider(names_from = category, values_from = avg_total_10_year) |>
+  # sum the total 10 year budget for each ward
+  group_by(ward_id, ward) |>
+  summarise(total_budget = sum(`Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`))
+
+# combine the model data with the ward profile data
+model_data <- model_data |>
+  left_join(analysis_ward_data, by = "ward_id") |>
+  select(ward_id, ward.x, total_budget, average_household_income, population) |>
+  rename(ward = ward.x) |>
+  # left join with the building permits data
+  left_join(analysis_building_permits, by = "ward_id") |>
+  select(ward_id, ward, total_budget, average_household_income, population, total_building_permits)
+
+posterior_vs_prior(model) + 
+  theme_classic() + 
+  theme(legend.position = "bottom") +
+  coord_flip()
+```
+
+## Markov Chain Monte Carlo Convergence
+
+Since `rstanarm` uses the MCMC sampling algorithm [@citeAlexander], we assess whether the algorithm encountered any issues by reviewing the Rhat and trace plots, as shown in [@fig-mcmc].
+The Rhat plot shows no concerns, with values consistently close to 1, indicating that the algorithm converged properly [@citeAlexander].
+Additionally, the trace plot appears normal, as the lines remain horizontal and fluctuate as expected, further supporting the adequacy of the sampling process [@citeAlexander].
+
+```{r}
+#| label: fig-mcmc
+#| fig-cap: "Checking the convergence of the MCMC algorithm"
+#| fig-subcap: ["Rhat plot", "Trace plot"]
+#| echo: false
+#| message: false
+#| warning: false
+#| fig-pos: H
+#| layout-ncol: 1
+#| fig-width: 10
+#| fig-height: 5
+
+plot(model, "rhat") + 
+  theme_classic()
+
+plot(model, "trace")
+
+```
+
+## Credibility Intervals
+
+Lastly, we employ a 95% credibility interval as showcased in @fig-ci, to gain a better understanding of the probability distribution of our coefficients.
+
+```{r}
+#| label: fig-ci
+#| fig-cap: "Credible intervals"
+#| echo: false
+#| message: false
+#| warning: false
+#| fig-pos: H
+
+plot(model) + 
+  theme_classic()
+
+```
+
+\newpage
 # References

### 2024-12-04 scripts/06-model_data.R

@@ -19,30 +19,75 @@ analysis_budget_data <- read.csv(here::here("data/02-analysis_data/analysis_budg
 analysis_building_permits <- read.csv(here::here("data/02-analysis_data/analysis_building_permits.csv"))
 analysis_ward_data <- read.csv(here::here("data/02-analysis_data/analysis_ward_data.csv"))
 
+
 ### Model data ####
 
-# combine all the years data for each ward and for each category
-model_data <- analysis_budget_data |>
-  group_by(ward_id, category, ward) |>
-  summarize(total_10_year = sum(total_10_year)) |>
-  pivot_wider(names_from = category, values_from = total_10_year)
-
-# combine the model data with the ward profile data
-model_data <- model_data |>
-  left_join(analysis_ward_data, by = "ward_id") |>
-  select(ward_id, ward.x, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
-  rename(ward = ward.x) |>
+model_data <-
+  analysis_budget_data |>
+  group_by(ward_id, ward, category) |>
+  summarise(total_10_year = mean(total_10_year)) |>
+  pivot_wider(names_from = category, values_from = total_10_year) |>
+  select(ward_id, ward, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
   left_join(analysis_building_permits, by = "ward_id") |>
-  rename(growth_related = `Growth Related`, state_of_good_repair = `State of Good Repair`, service_improvement = `Service Improvement and Enhancement`)
+  left_join(analysis_ward_data, by = "ward_id") |>
+  rename(ward = ward.x) |>
+  select(ward_id, ward, population, average_household_income, total_building_permits, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
+  # add a new column for total budget across three categories
+  mutate(total_budget = `Growth Related` + `State of Good Repair` + `Service Improvement and Enhancement`) |>
+  select(-`Growth Related`, -`State of Good Repair`, -`Service Improvement and Enhancement`)
 
+# mean of all population, average household income, total budget and total building permits
+mean_population <- mean(model_data$population)
+mean_income <- mean(model_data$average_household_income)
+mean_budget <- mean(model_data$total_budget)
+mean_building_permits <- mean(model_data$total_building_permits)
+
+# median of all population, average household income, total budget and total building permits
+median_population <- median(model_data$population)
[…]

### 2024-12-04 other/datasheet/datasheet_template.qmd
Updated Files
- Added Datasheet
- Added LLM usage
- Final checks in paper
- Updated reference.bib

---
title: "Datasheet for Open Data Toronto Ward Budget Dataset"  
subtitle: "Comprehensive Overview of Ward Budgets (2021-2024)"  
author:  
  - Aviral Bhardwaj  
thanks: "Code and data are available at: [https://github.com/Aviral-03/InfrastructureCausalModel-TO](https://github.com/Aviral-03/InfrastructureCausalModel-TO)."  
date: today
date-format: long 
format: pdf  
number-sections: true  
bibliography: references.bib  
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The dataset was created to provide a comprehensive overview of the Ward Budgets in Toronto from 2021-2024. It aims to offer detailed insights into budget allocation and spending patterns across different wards, helping identify potential disparities or inequities in resource allocation.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - The dataset was created by Aviral Bhardwaj, with the source being Open Data Toronto [@toronto_budget_plan]
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - The project is self-funded by the creator, with no external grant support.
4. *Any other comments?*
    - There are no additional comments.

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- The dataset represents budget allocations for Toronto wards. It includes multiple instance types such as wards, budget categories, budget amounts, and years (2021-2024).
2. *How many instances are there in total (of each type, if appropriate)?*
	- The dataset contains 25 wards in Toronto, covering 4 years (2021-2024), for a total of 100 ward-year combinations.
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- The dataset appears to be a comprehensive collection of ward budgets, covering all 25 wards in Toronto for the specified time period. It represents official budget data from Open Data Toronto [@toronto_budget_plan]
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- Raw data contains these headers: Program/Agency Name, Project Name, Sub-Project Name, Total 10 Year, Ward Number, Ward, Category
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- Instances are labeled by ward identifier, budget category, year, and budget amount.
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- No information is missing from the instances.
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- Instances are related through consistent ward identifiers, comparable budget categories across years, and geographical relationships between wards.
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- The data can be naturally split by ward, year, and budget category.
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- Potential considerations include budget rounding, categorization inconsistencies, and currency fluctuations. It is recommended to cross-reference with official sources.
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- The dataset is primarily based on Open Data Toronto [@toronto_budget_plan] and recommends keeping the original source documentation. There are no known external resource restrictions.
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- The dataset contains public budget data with no personally identifiable information. Data is aggregated at the ward level.
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- The dataset contains no offensive or anxiety-inducing content, as it is purely factual budget information.
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- The dataset is divided by ward, covering all 25 Toronto wards, but it does not provide demographic breakdowns within the wards.
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- There is no direct individual identification within the dataset, as it is aggregated at the ward level.
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- The dataset contains public financial data and no sensitive personal information. Budget allocations are public records.
16. *Any other comments?*
	- No additional comments.

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- The data was directly sourced from Open Data Toronto [@toronto_budget_plan], an official government data portal.
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- The dataset was likely obtained through a software API or downloaded from Open Data Toronto [@toronto_budget_plan], with manual compilation and verification for accuracy.
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- The dataset provides a comprehensive collection, including all 25 Toronto wards and covering all years from 2021 to 2024.
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- The data was collected by Open Data Toronto [@toronto_budget_plan] and was made available by Government of Toronto.
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- The budget planning and allocation data was collected over the years 2021-2024, matching the creation timeframe of the dataset. Each year on February 1st, the City of Toronto releases the budget for the upcoming year.
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- The dataset follows ethical guidelines, as it consists of publicly available data with no personal privacy concerns, adhering to open government data principles.
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- The data was obtained from Open Data Toronto, an official government data portal.
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- N.A
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- Public budget data is not subject to individual consent requirements, as it is publicly available information.
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- N.A
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- Public budget data does not require a data protection impact analysis, as it is publicly available information.
12. *Any other comments?*
	- N.A

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- Potential preprocessing steps include standardizing budget category names, ensuring consistent formatting, and converting data to appropriate types, and combining each year's data into a single dataset.
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- The raw data is available on Open Data Toronto, and the preprocessed data is available in the dataset. Link: [Open Data Toronto](https://open.toronto.ca/dataset/budget-capital-budget-plan-by-ward-10-yr-approved/) 
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- `opendatatoronto` package in R was used to download the data. The preprocessing steps were done in R and Python.
4. *Any other comments?*
	- N.A

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- The dataset can be used for urban policy analysis, budget allocation research, comparative ward-level budget studies, and local government funding analysis.
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- No
3. *What (other) tasks could the dataset be used for?*
	- TBD
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- The dataset is based on official government records and should be used responsibly, avoiding any unfair treatment or misrepresentation of the data. It is recommended to cross-reference with official sources and consider the context of budget allocation.
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- No.
6. *Any other comments?*
	- No.

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- The dataset is likely to be shared for research or educational purposes, with no commercial distribution intended.
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- The dataset can be shared via platforms such as GitHub or a research repository. It is also recommended to assign a DOI for academic citation.
3. *When will the dataset be distributed?*
	- The dataset is available now and can be distributed upon request.
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- The dataset follows Open Data Toronto's licensing terms, with a recommendation to use a Creative Commons Attribution license and to cite the original source.
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- No third-party restrictions are known to be associated with the dataset.
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- No export controls or regulatory restrictions are known to apply to the dataset.
7. *Any other comments?*
	- No

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- The dataset is maintained by Aviral Bhardwaj, and annual updates are recommended and raw data is available on Open Data Toronto [@toronto_budget_plan].
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- Contact information has been provided in the author section.
3. *Is there an erratum? If so, please provide a link or other access point.*
	- No.
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- It is suggested that the dataset be reviewed annually, with updates to include new budget years and verification of data accuracy. Updates will be communicated through the dataset repository.
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- No personal data is included in the dataset, so there are no retention limits.
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- No
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- Contributions are welcome, with a suggested peer review process and validation against official sources.
8. *Any other comments?*
	- No.

\newpage

# References

### 2024-12-04 paper/paper.qmd
Updated Files
- Added Datasheet
- Added LLM usage
- Final checks in paper
- Updated reference.bib

@@
   - Aviral Bhardwaj
-thanks: "Code and data are available at: https://github.com/Aviral-03/Toronto-WardWide-Budget-Analysis"
+thanks: "Code and data are available at: [https://github.com/Aviral-03/InfrastructureCausalModel-TO](https://github.com/Aviral-03/InfrastructureCausalModel-TO)." 
 date: today
@@
 Using data from the city's \"Capital Budget Plan\", \"Ward Profiles\" and \"Active Building Permits\", we analyze spending patterns in key areas like related to infrastructre.
-Our analysis, guided by a causal model, reveals a negative association between population density and construction activity, with some high-density wards receiving significantly higher budget allocations.
+Our analysis, guided by a causal model, presents a negative association between population density and construction activity, with some high-density wards receiving significantly higher budget allocations.
 Average household income and total budget allocations were positively correlated with the number of building permits.
-These insights highlight potential disparities in resource distribution, emphasizing the need for more equitable urban budget planning to address diverse community needs."
+These findings highlight potential disparities in resource distribution, emphasizing the need for more equitable urban budget planning to address diverse community needs."
 format: pdf
@@
 [@sec-discussion] explores the implications for urban governance and budget planning in Toronto.
-Finally, [@sec-limitations] addresses the study’s limitations and [@sec-future-work] offers recommendations for future research.
+Finally, [@sec-limitations] addresses the study’s limitations and offers recommendations for future research.
 Additional details, including tables, figures, and methodology, are provided in the [@sec-appendix].
@@
 
-Population is a critical demographic indicator, representing the number of residents in each ward, while average household income reflects economic well-being, influencing access to resources and quality of life [@Schaeffer].
+Population is a important demographic indicator, representing the number of residents in each ward, while average household income reflects economic well-being, influencing access to resources and quality of life [@Schaeffer].
 Therefore, these two variables of interest were chosen for the analysis.
@@
 
-This data-set was included in this analysis to provide insights into the population and average household income for each ward, providing insights into the city's socioeconomic landscape.
+This data-set was included in this analysis to provide understanding into the population and average household income for each ward, providing understanding into the city's socioeconomic landscape.
 25-Ward model was used instead of the 44-Ward model as it was the most recent data available at the time of analysis and matched the Capital Budget data.
@@
 
-These categories were selected because they represent critical areas of investment that directly impact the quality of life and well-being of Toronto residents.
+These categories were selected because they represent significant areas of investment that directly impact the quality of life and well-being of Toronto residents.
 *State of Good Repair* focuses on maintaining and preserving existing infrastructure, *Growth Related* addresses the expansion and development of new infrastructure, and *Service Improvement and Enhancement* aims to enhance public services and amenities.
-By analyzing budget allocations in these categories, we can gain valuable insights into the city's spending priorities and identify opportunities for more effective and equitable resource distribution.
+By analyzing budget allocations in these categories, we can gain understand city's spending priorities and identify opportunities for more effective and equitable resource distribution.
 While also identifying potential disparities in funding across various wards of the city.
@@
 
-The *Building Permits - Active Permits* dataset [@toronto_building_permits], published by the City Planning Division and last updated on November 28, 2024, serves as a comprehensive record of active building permits in Toronto.
+The *Building Permits - Active Permits* dataset [@toronto_building_permits], published by the City Planning Division and last updated on November 28, 2024, serves as an important record of active building permits in Toronto.
 A building permit is a municipally issued document, mandated by the Building Code Act and enforced by the City of Toronto, regulating the construction or demolition of physical structures [@toronto_building_permits].
@@
 The number of permits serves as a proxy for construction activity and development, highlighting the demand for infrastructure investment and capital expenditures.
-By examining the relationship between building permits and budget allocations, we can gain valuable insights into how construction activity influences resource distribution across the city's wards.
+By examining the relationship between building permits and budget allocations, we can understand how construction activity influences resource distribution across the city's wards.
 
@@
 
-[@fig-building-permits-data] illustrates the total number of building permits by ward, providing valuable insights into construction activity and development patterns across Toronto’s wards.
+[@fig-building-permits-data] illustrates the total number of building permits by ward, providing understanding into construction activity and development patterns across Toronto’s wards.
 
@@
 ```
-### Relationship between Average Household Income & Population with Budget Allocation for Service Improvement and Enhancement Projects 
+
+### Relationship between Predictors and Budget Allocations (Service Improvement and Enhancement)
 ```{r}
@@
 
-### Relationship between Average Household Income & Population with Budget Allocation for Growth Related Projects and State of Good Repair Projects
-
-[@fig-growth-related-budget] (a) illustrates the relationship between average household income, population, and the total 10-year budget allocation for Growth-Related projects.
+### Relationship between Predictors and Budget Allocations (Growth-Related and State of Good Repair)
+
+[@fig-growth-related-budget]a) illustrates the relationship between average household income, population, and the total 10-year budget allocation for Growth-Related projects.
 The linear regression line for average household income remains relatively flat, hovering around \$100,000.
 This suggests that budget allocations for Growth-Related projects are not significantly influenced by income levels, indicating a more uniform distribution of funds across wards regardless of household income.
-A similar lack of correlation is observed in [@fig-growth-related-budget] (b), where budget allocations for State of Good Repair projects also show minimal dependence on average household income.
+A similar lack of correlation is observed in [@fig-growth-related-budget]b), where budget allocations for State of Good Repair projects also show minimal dependence on average household income.
 
@@
 Wards with higher population densities—such as Ward 10 (Spadina-Fort York), Ward 13 (Toronto Centre), Ward 5 (York South-Weston), and Ward 2 (Etobicoke North)—receive a larger share of the Growth-Related budget.
-This indicates that population density plays a critical role in determining funding levels for these projects, with more populous wards benefiting from increased allocations.
+This indicates that population density plays an important role in determining funding levels for these projects, with more populous wards benefiting from increased allocations.
 A similar trend is observed in the budget distribution for State of Good Repair projects, where wards with higher population densities receive a proportionally larger share of the funds.
+
 ```{r}
@@
 #| label: fig-growth-related-budget
-#| fig.cap: "Growth Related Budget and State of Good Repair Budget by Ward"
-#| fig-subcap: ["Growth Related Budget", "State of Good Repair Budget"]
+#| fig.cap: "Growth and State of Good Repair Budget by Ward"
+#| fig-subcap: ["Growth Related", "State of Good Repair"]
 #| fig-pos: "h"
 #| fig.width: 10
-#| fig.height: 5
-#| layout-ncol: 1
+#| fig.height: 3
 
@@
   theme_minimal()
-
-# Combine both plots side by side
-grid.arrange(p1, p2, ncol = 2)
-
 good_repair_budget <- combined_data |>
@@
 
-p1 <- ggplot(good_repair_budget, aes(x = population, y = `State of Good Repair`, label= ward_id)) +
+p3 <- ggplot(good_repair_budget, aes(x = population, y = `State of Good Repair`, label= ward_id)) +
   geom_point() +
@@
 
-p2 <- ggplot(good_repair_budget, aes(x = average_household_income, y = `State of Good Repair`, label= ward_id)) +
+p4 <- ggplot(good_repair_budget, aes(x = average_household_income, y = `State of Good Repair`, label= ward_id)) +
   geom_point() +
@@
 grid.arrange(p1, p2, ncol = 2)
-
-```
-
-\newpage
+grid.arrange(p3, p4, ncol = 2)
+
+# | echo: false
+# | message: false
+# | warning: false
+# | label: fig-good-repair-budget
+# | fig.cap: "State of Good Repair Budget by Ward"
+# | fig-pos: "h"
+# | fig.width: 10
+# | fig.height: 5
+```
 
@@
 
-To better understand the underlying data, [@eda-model-justification] provides a summary of the mean, median, and standard deviation for each variable used in the model.
-The predictor variables have large values and different scales, which can cause numerical instability and convergence issues in the model.
-To address this, we will standardize the predictor variables by subtracting the mean and dividing by the standard deviation.
-This will scale the variables consistently and improve both the model's convergence and interpretability.
-Additionally, we did not use a Bayesian Poisson regression model or negative binomial because the number of building permits is a continuous variable, not count data.
-Therefore, we chose to use a Bayesian multiple linear regression model.
-This model allows us to estimate the effect of population density, average household income, and budget allocations on the number of building permits issued in each ward giving us a well-rounded understanding of the relationships between these variables.
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: eda-model-justification
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: tbl-model-justification
 #| tbl-cap: "Mean, Median, and Standard Deviation of Each Variable"
@@
 
+To better understand the underlying data, [@tbl-model-justification] provides a summary of the mean, median, and standard deviation for each variable used in the model.
+The predictor variables have large values and different scales, which can cause numerical instability and convergence issues in the model.
+To address this, we will standardize the predictor variables by subtracting the mean and dividing by the standard deviation.
+This will scale the variables consistently and improve both the model's convergence and interpretability.
+Additionally, we did not use a Bayesian Poisson regression model or negative binomial because the number of building permits is a continuous variable, not count data.
+Therefore, we chose to use a Bayesian multiple linear regression model.
+This model allows us to estimate the effect of population density, average household income, and budget allocations on the number of building permits issued in each ward giving us a well-rounded understanding of the relationships between these variables.
+
 The factors chosen for the model are known to influence the number of building permits issued in each ward.
@@
 
-Population distribution reveals a concentration in the western part of the city and downtown areas.
+Population distribution presents a concentration in the western part of the city and downtown areas.
 Notably, wards like Etobicoke Centre (Ward 2) and Etobicoke-Lakeshore (Ward 3), situated farther from downtown, have lower costs of living and correspondingly lower average household incomes, such as Etobicoke North (Ward 2) with \$95,200.
@@
 
-The results of the Bayesian regression model, summarized in [@tbl-model-results], provide important insights into the relationship between total budget allocations, average household income, and the number of building permits issued across wards.
-
-The model explains 59.1% of the variance in building permits (R² = 0.591), indicating a moderate level of explanatory power.
+The results of the Bayesian regression model, summarized in [@tbl-model-results], provide important understanding into the relationship between total budget allocations, average household income, and the number of building permits issued across wards.
+
+The model explains 59.1% of the variance in building permits, indicating a moderate level of explanatory power.
 The coefficients suggest that both total budget (0.530) and average household income (0.350) have a positive and significant impact on the number of building permits issued, implying that areas with higher budget allocations and income levels are more likely to see increased construction activity.
 
-The inclusion of ward-specific random effects (σ² = 0.141) highlights the importance of accounting for variation between wards, suggesting that there are unobserved factors specific to each ward influencing permit issuance.
-Although the adjusted R² (0.369) indicates that the model's explanatory power is somewhat limited after accounting for the complexity of the predictors, the model still provides valuable insights into the role of economic factors in shaping construction activity.
+The inclusion of ward-specific random effects highlights the importance of accounting for variation between wards, suggesting that there are unobserved factors specific to each ward influencing permit issuance.
+Although the adjusted $R^2$ (0.369) indicates that the model's explanatory power is somewhat limited after accounting for the complexity of the predictors, the model still provides understanding into the role of economic factors in shaping construction activity.
 With an RMSE of 0.58, the model's predictions are reasonably accurate, although there is still room for improvement.
@@
 
-As highlighted in [@sec-model-results] Bayesian regression model reveals that average household income has a significant positive effect on the issuance of building permits.
+As highlighted in [@sec-model-results] Bayesian regression model shows that average household income has a significant positive effect on the issuance of building permits.
 This suggests that construction activity tends to concentrate in higher-income areas, where demand for development is stronger.
@@
 
-For potential property investors in Toronto, these insights offer a strategic advantage.
+For potential property investors in Toronto, these understanding offer a strategic advantage.
 Investing in areas with higher average household incomes may be more profitable due to the increased likelihood of ongoing construction activity, which can drive property value appreciation and strengthen the local economy.
@@
 
-These findings underscore the importance of equity-based budgeting, where low-density and lower-income areas should receive targeted investments to meet critical infrastructure needs.
+These findings underscore the importance of equity-based budgeting, where low-density and lower-income areas should receive targeted investments to meet infrastructure needs.
 This approach aligns with urban development strategies that prioritize inclusivity.
@@
 
-# Weaknesses and Limitations {#sec-limitations}
+# Weaknesses and Future Work {#sec-limitations}
 
@@
 As discussed in [@sec-data], the datasets used in this study are publicly available from the City of Toronto.
-While they provide valuable insights into the city's demographics, budget allocations, and building permits, they may not fully capture the complexity of Toronto’s infrastructure development and resource distribution.
+While they represents city's demographics, budget allocations, and building permits, they may not fully capture the complexity of Toronto’s infrastructure development and resource distribution.
 For example, the building permits dataset includes only active permits, which may not reflect the entire scope of construction activity.
@@
 This analysis focused on only one subset of these statuses, which may not fully capture the complexity of the building permit process and the various stages of construction.
-Future research could include additional permit statuses for a more comprehensive understanding of construction activity.
+Future research could include additional permit statuses for a better understanding of construction activity.
 Another limitation is the lack of detailed data on specific projects within each budget category.
-This information could offer further insights into why certain wards receive more or less funding.
+This information could offer further understanding into why certain wards receive more or less funding.
 
@@
 
-# Future Work {#sec-future-work}
+## Future Work
 
@@
 Future studies could explore additional factors that influence budget allocations and building permit issuance, such as infrastructure needs, community priorities, and political considerations.
-By incorporating these variables into the analysis, researchers can gain a more comprehensive understanding of how municipal budgets are allocated and how construction activity is distributed across the city's wards.
+By incorporating these variables into the analysis, researchers can gain a more in-depth understanding of how municipal budgets are allocated and how construction activity is distributed across the city's wards.
 
@@
 
+\appendix
+
 # Appendix {#sec-appendix}
+
+## Survey, Sample, and Data Collection
+
+The relationship between income levels, population and construction outcomes is complex and influenced by diverse socioeconomic and regional factors.
+Current observational data, such as NHS survey to get income levels, and population of ward profiles but is inherently limited by its sample [@nhs_canada].
+To better answer our research question, a more targeted survey approach is essential.
+This exploration outlines an idealized survey methodology designed to reduce bias, ensure representativeness, and deepen understanding.
+
+### Idealized Survey Design for Construction Analysis 
+A survey specifically designed to explore the link between income and construction outcomes would involve the following key components:
+
+a. Defining the Survey Objectives
+The survey would aim to investigate how income disparities influence construction investment, access to resources, regulatory hurdles, and housing quality across different regions.
+
+b. Sampling Methodology  
+To ensure a representative and unbiased dataset, the survey would employ:
+
+- Stratified Sampling: Income levels and ward classifications (e.g., urban, suburban, rural) would be stratified to ensure proportional representation.  
+- Random Sampling Within Strata: Participants would be randomly selected within each stratum to enhance the validity and generalizability of the findings.
+
+c. Sample Size and Distribution
+The sample size should be sufficiently large to allow for statistically significant comparisons across different income levels and geographic regions.
+A minimum target of several thousand participants distributed evenly across strata would be ideal.
+
+### Reducing Bias in Sampling and Data Collection  
+Bias reduction is important in survey design to ensure reliable and actionable understanding. This involves addressing various types of bias:
+
+a. Sampling Bias Mitigation
+Stratified random sampling ensures that underrepresented groups, particularly those in lower income brackets or remote regions, are adequately included.
+
+b. Non-Response Bias Mitigation 
+Offering multiple survey delivery methods—such as online platforms, in-person interviews, and mailed surveys—enhances accessibility and participation.
+Additionally, providing incentives like small monetary rewards or gift cards can boost response rates among diverse groups.
+
+c. Design Bias Mitigation 
+To avoid influencing responses, survey questions must be neutrally worded and pilot-tested to ensure clarity.
+Cognitive interviews with a subset of participants can help refine the wording and structure of questions.
 

### 2024-12-04 scripts/05-model_data.R
Updated Scripts

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Model data for the analysis
+# Author: Aviral Bhardwaj
+# Date: 2024-12-01
+# Contact: aviral.bhardwaj@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: N.A
+# Any other information needed? N.A
 
 
 #### Workspace setup ####

### 2024-12-14 paper/paper.qmd
Final Paper Edits

@@
 title: "Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto"
-subtitle: "Population Density Drive Larger Budget Allocations meanwhile Average Household Income Drives Construction Projects"
+subtitle: "Population density drive larger budget allocations meanwhile average household income drives construction projects"
 author: 
@@
 date-format: long
-abstract: "This study investigates how population density and average household income affect budget allocations across Toronto’s 25 wards, and how these factors, along with budget allocation, influence construction projects.
-Using data from the city's \"Capital Budget Plan\", \"Ward Profiles\" and \"Active Building Permits\", we analyze spending patterns in key areas like related to infrastructre.
-Our analysis, guided by a causal model, presents a negative association between population density and construction activity, with some high-density wards receiving significantly higher budget allocations.
-Average household income and total budget allocations were positively correlated with the number of building permits.
-These findings highlight potential disparities in resource distribution, emphasizing the need for more equitable urban budget planning to address diverse community needs."
+abstract: "This study examines the relationship between population density, average household income, and budget allocations across Toronto's 25 wards and investigates their combined influence on construction activity.
+Using data from the city's *Capital Budget Plan*, *Ward Profiles*, and *Active Building Permits*, we analyze spending patterns in key infrastructure-related areas.
+Employing a causal modeling approach, our findings reveal a negative association between population density and construction activity, despite some high-density wards receiving higher budget allocations.
+Additionally, average household income and total budget allocations exhibit a positive correlation with the number of building permits.
+These results underscore potential inequities in resource distribution, highlighting the need for data-driven and equitable urban budget planning to better address community needs."
 format: pdf
@@
 In urban governance, the equitable allocation of municipal budgets is essential to meet a city’s social, economic, and infrastructure needs.
-This challenge is especially pressing in Toronto, where 25 wards compete for limited resources amid a growing population.
-Understanding how demographic factors like population and income influence budget decisions remains underexplored.
-This study seeks to answer: How do population and income affect budget allocations across Toronto’s wards, and do these factors, along with budget allocation, impact the number of construction projects, a key indicator of infrastructure development?
-
-Historical budget data shows uneven investment, particularly in high-density, lower-income areas, leading to disparities in infrastructure and services [@ferguson2024compromised].
-Following her 2023 election, Mayor Olivia Chow prioritized housing, transit, and equity-focused initiatives [@toronto_2024_budget].
-The 2024 budget reflects these priorities, with significant investments in affordable housing, transportation, and community safety, while emphasizing long-term infrastructure development through a 10-year capital plan.
-However, it remains unclear if these allocations are equitably distributed across wards or align with Toronto’s broader economic and demographic realities.
-
-Major projects like the Ontario Line—a new subway connecting Toronto’s east and west ends—are expected to spur economic growth and development [@citeTransitExpansion].
-Similarly, Olivia Chow’s "Sidewalks to Skylines" 10-year plan aims to create jobs and improve residents' quality of life [@citeSidewalksToSkylines].
-As these initiatives unfold, equitable infrastructure investments will be vital for Toronto’s long-term prosperity.
-With more residents expected to spread across the city's wards, it is essential to evaluate whether current budget allocations align with the city’s broader development goals as the demographic characteristics of each ward changes.
-
-This study uses data from the City of Toronto’s Open Data Portal, including demographic information from the "2021 Ward Profiles", "Capital Budget & Plan By Ward (10 yr Approved) 2021-2024", and "Building Applications and Permits".
-Our analysis, guided by a causal model, shows that both population density and average household income significantly influence budget allocations and the issuance of building permits, key indicators of infrastructure development.
-However, population density was identified as an exclusion restriction, as its inclusion in the Bayesian multiregression model revealed a counterintuitive negative relationship with the number of building permits.
-This suggests that unobserved factors may be influencing the link between population density and construction activity.
-After excluding population density from the model, we found that average household income and total budget allocations were positively correlated with building permits.
-This result aligns with expectations: wealthier wards with higher incomes receive larger budget allocations, leading to more construction projects.
+This challenge is especially pressing in Toronto, where 25 wards compete for limited resources amid a growing population [@Ontario2024DemographicTrends].
+Understanding how demographic factors influence budget decisions remains under-explored.
+Therefore, this study seeks to answer: How do demographic factors such as population and income affect budget allocations across Toronto’s wards, and can we assess their combined impact on construction activity, a key indicator of infrastructure development?
+
+Looking back, Toronto's historical budget data have shown uneven investment, particularly in high-density, lower-income areas, leading to disparities in infrastructure and services [@ferguson2024compromised].
+However, following 2023 Mayor election, Mayor Olivia Chow promise to prioritize housing, transit, and equity-focused initiatives [@toronto_2024_budget].
+The 2024 budget prioritizes significant investments in affordable housing, transportation, and community safety, while also focusing on long-term infrastructure development through a 10-year capital plan.
+Notable projects, such as the *Ontario Line* subway expansion, are projected to drive economic growth and urban development [@citeTransitExpansion].
+Similarly, the *Sidewalks to Skylines* 10-year plan aims to create jobs and enhance residents' quality of life [@citeSidewalksToSkylines].
+As these initiatives progress, ensuring equitable infrastructure investments will be crucial for Toronto’s sustainable growth.
+With the city’s population increasingly distributed across its wards, it is imperative to assess whether current budget allocations effectively align with broader development objectives amidst shifting demographic trends.
+
+This study analyzes data from the City of Toronto’s Open Data Portal, incorporating demographic information from the *2021 Ward Profiles*, budget details from the *Capital Budget & Plan By Ward (10-Year Approved) 2021-2024*, and construction data from *Building Applications and Permits*.
+Using a Bayesian multi-regression model guided by a causal framework, we found that both average household income and total budget allocations significantly influence the number of building permits.
+However, population density displayed an unexpected negative relationship with construction activity, suggesting the presence of unobserved factors affecting this link.
+To address this, we excluded population density as an explanatory variable and confirmed that higher-income wards receiving larger budget allocations corresponded to more construction activity.
+These findings highlight the importance of understanding how socioeconomic and budgetary factors shape urban development, emphasizing the need for equitable resource distribution to support balanced growth across Toronto’s wards.
 
 The paper is organized as follows: [@sec-data] outlines the data sources and methodology used in the analysis.
-[@sec-model] explains the Bayesian model applied to estimate the effects of demographic and economic factors, along with budget allocations, on construction projects.
+[@sec-model] explains the Bayesian model applied to estimate the effects of demographic and economic factors, following the causal model ([@fig-causal-model]).
 [@sec-results] presents key findings, including trends in budget allocations, the relationship between demographic factors and budget distribution, and the impact on construction projects.
@@
 
-Our research question and estimand analyse the causal relationships between population, income, budget allocation, and active construction projects ([@fig-causal-model]).
-Population and income serve as confounders, influencing both budget allocation and construction projects.
-Budget allocation acts as a mediator, connecting the confounders to the outcome, and as an instrumental variable, helping to estimate the effect of these predictors on the number of construction projects.
-The outcome variable is active construction projects, while population and income are the predictor variables.
-We aim to estimate the causal effect (estimand) of budget allocation on construction projects, while accounting for the influence of population and income.
-This paper highlights how budget decisions mediate the broader impact of socioeconomic factors on infrastructure development.
+Our research investigates the causal relationships between population, income, budget allocation, and active construction projects, as illustrated in [@fig-causal-model].
+Population and income are key confounders, influencing both budget allocation and the number of construction projects.
+Budget allocation serves a dual role: as a mediator connecting population and income to construction projects and as an instrumental variable, helping to estimate their effects on construction activity.
+The outcome variable is the number of active construction projects, while population and income are the primary predictors.
+Our goal is to estimate the causal effect of budget allocation on construction activity, accounting for the confounding effects of population and income.
 
@@
 #| label: fig-causal-model
-#| fig-cap: "Causal model"
-#| fig-subcap: "Population and income influence budget allocation and construction projects"
+#| fig-cap: "Causal model: Population and income influence budget allocation and construction projects"
 #| fig-pos: "h"
+#| fig.align: "center"
 
@@
 The `readxl` [@readxl] package was used for reading Excel files.
-Other R packages used which includes `tidyverse` [@tidyverse], `styler` [@styler], and `dplyr` [@dplyr] for creating tables.
-The `ggplot2` [@ggplot2] and `kableExtra` [@kableExtra] were used for data visualization and table formatting.
+Other R packages used includes `tidyverse` [@tidyverse], `styler` [@styler], and `dplyr` [@dplyr] for creating tables.
+`ggplot2` [@ggplot2] and `kableExtra` [@kableExtra] were used for data visualization and table formatting.
 The `patchwork` [@patchwork] package was used for combining multiple plots, and `sf` [@sf] for spatial data analysis.
+For model estimation, the `rstanarm` [@citerstanarm] package was used for Bayesian modeling, and `bayesplot` [@citeBayesPlot] for visualizing the results.
 The `lintr` [@citeLintr] package was used for code linting.
@@
 
-These categories were selected because they represent significant areas of investment that directly impact the quality of life and well-being of Toronto residents.
+Since our analysis aims to understand how budget allocations influence construction activity and infrastructure development, we selected these three categories as they are most relevant to capital expenditures and infrastructure projects.
 *State of Good Repair* focuses on maintaining and preserving existing infrastructure, *Growth Related* addresses the expansion and development of new infrastructure, and *Service Improvement and Enhancement* aims to enhance public services and amenities.
-By analyzing budget allocations in these categories, we can gain understand city's spending priorities and identify opportunities for more effective and equitable resource distribution.
-While also identifying potential disparities in funding across various wards of the city.
 
@@
 The City Wards dataset [@toronto_city_wards], published by the City Clerk's Office and last updated on July 22, 2024, contains geographical information about each ward, including the ward ID, ward name, and ward boundary.
-These ward boundaries were decided as a part of `Bill 5, Better Local Government Act` in 2018, reducing the number of wards from 47 to 25 [@toronto_city_wards].
+These ward boundaries were decided as a part of *Bill 5, Better Local Government Act* in 2018, reducing the number of wards from 47 to 25 [@toronto_city_wards].
 
@@
 The Building Division reviews the application, and a Toronto Building Inspector ensures compliance with the Ontario Building Code, Zoning By-law, and other applicable regulations.
-Once approved, the permit is issued, allowing the applicant to commence construction or demolition.
+Once approved, the permit is issued, allowing the applicant to commence construction or demolition, and thus data is saved in the system as an active permit.
 
@@
 
-We selected this data-set to evaluate how budget allocations correlate with the number of building permits issued in each ward.
+We selected this data-set to evaluate how our confounders, population and income, influence construction activity, and mediator, budget allocation, correlate with the number of building permits issued in each ward.
 The number of permits serves as a proxy for construction activity and development, highlighting the demand for infrastructure investment and capital expenditures.
-By examining the relationship between building permits and budget allocations, we can understand how construction activity influences resource distribution across the city's wards.
+By examining the relationship between building permits and outlined variables, we can understand how construction activity influences resource distribution across the city's wards.
 
@@
 
-There are several limitations to the analysis that need to be addressed.
-As discussed in [@sec-data], the datasets used in this study are publicly available from the City of Toronto.
-While they represents city's demographics, budget allocations, and building permits, they may not fully capture the complexity of Toronto’s infrastructure development and resource distribution.
-For example, the building permits dataset includes only active permits, which may not reflect the entire scope of construction activity.
-Additionally, the budget allocations dataset may exclude certain infrastructure projects and expenditures, leading to potential gaps in the analysis.
-
-Furthermore, as detailed in [@sec-appendix], there are 45 unique statuses assigned to each submitted permit.
-This analysis focused on only one subset of these statuses, which may not fully capture the complexity of the building permit process and the various stages of construction.
-Future research could include additional permit statuses for a better understanding of construction activity.
-Another limitation is the lack of detailed data on specific projects within each budget category.
-This information could offer further understanding into why certain wards receive more or less funding.
-
-Lastly, the model assumes that the relationships between population density, average household income, and budget allocations are linear, which may oversimplify these relationships.
-Additionally, the analysis assumes independence among wards, which may not be accurate due to shared resources and infrastructure projects.
-Future research could explore more advanced modeling techniques to address these non-linear relationships and dependencies.
+This analysis has several limitations that warrant consideration.
+As discussed in [@sec-data], the datasets used, sourced from the City of Toronto, provide insights into the city’s demographics, budget allocations, and building permits.
+However, they may not fully capture the complexity of Toronto's infrastructure development and resource distribution.
+
+For instance, in [@sec-building-permits], all building *Work Types* were included in the analysis, which may not accurately represent construction activity in each ward.
+Many recorded work types, such as renovations, demolitions, and minor repairs (e.g., HVAC installations or pipework), are not indicative of significant construction projects.
+Future research could refine the analysis by focusing on specific types of construction projects to better reflect actual development activity.
+
+Additionally, as noted in [@sec-appendix], there are 45 unique statuses assigned to building permits, but this study analyzed only a subset of these statuses.
+This narrow focus may overlook important stages in the construction process.
+Expanding the analysis to include more permit statuses could offer a more comprehensive understanding of construction activity.
+
+Another limitation is the lack of detailed data on specific projects within each budget category, which restricts insights into why certain wards receive more or less funding.
+Access to granular project-level data could provide deeper context and explanations for funding disparities.
+
+Moreover, the analysis assumes linear relationships between population density, average household income, and budget allocations, which may oversimplify the underlying dynamics.
+Developing more complex models that account for non-linear relationships and interactions could yield a richer understanding of factors influencing construction activity.
+
+Finally, the analysis assumes independence among wards, an assumption that may not hold true given shared infrastructure and resource dependencies.
+Future studies could incorporate spatial or network-based models to address interdependencies between wards.
 

### 2024-12-15 paper/paper.qmd
Final Commit
- Paper edits finished

@@
 Ward 10 (Spadina-Fort York) received the highest budget allocations across both *Growth Related* and *State of Good Repair* categories, indicating significant investment in infrastructure development and maintenance to population-dense areas.
-Toronto Center, Don Valley East, and Etobicoke Center also received notable budget allocations, some of these wards have higher average household incomes as well as higher population densities.
+Toronto Center (Ward 13), Don Valley East (Ward 16), and Etobicoke Center (Ward 2) also received notable budget allocations, some of these wards have higher average household incomes as well as higher population densities.
 The box plots reveal that budget allocations for *Growth Related* projects have increased steadily over the years, with 75% of wards receiving higher allocations in 2024 compared to 2021.
@@
 
-### Relationship between Predictors and Budget Allocations (Service Improvement and Enhancement)
+### Relationship between Confounders and Budget Allocations (Service Improvement and Enhancement)
 ```{r}
@@
 ```
-[@fig-service-improvement-budget] illustrates the relationship between average household income, population, and the total 10-year budget allocation for Service Improvement and Enhancement projects.
+[@fig-service-improvement-budget] illustrates the relationship between average household income, population, and the total budget allocation each year (2021-2024) for *Service Improvement and Enhancement* projects.
 These projects  aim to enhance the quality of services provided to residents and improve the overall infrastructure of the city [@toronto_improvements].
 
-Unlike Growth-Related and State of Good Repair projects, the linear regression lines for average household income and population shows a positive correlation with budget allocation for Service Improvement and Enhancement projects.
-Wards with higher average household incomes, such as Ward 11 (University-Rosedale) and Ward 13 (Toronto Centre), receive a larger share of the budget allocation for these projects.
-This suggests that higher-income areas are prioritized for service improvements and infrastructure enhancements, reflecting a targeted approach to resource allocation based on income levels.
-
-Other wards like Ward 3 (Etobicoke-Lakeshore) and Ward 10 (Spadina-Fort York), which have high population densities, also received significant budget allocations, suggesting that these areas are priorities for service improvements and infrastructure enhancement.
+The linear regression lines for average household income and population shows a positive correlation with budget allocation for Service Improvement and Enhancement projects.
+Wards with higher average household incomes, such as Ward 11 (University-Rosedale) receive a larger share of the budget allocation for these projects, while wards with higher population densities, such as Ward 10 (Spadina-Fort York) and Ward 3 (Etobicoke-Lakeshore), also receive a proportionally larger share of the funds.
+This suggests that both income levels and population density play a significant role in determining funding levels for Service Improvement and Enhancement projects, with higher-income and more populous wards benefiting from increased allocations.
+
+Other wards like Ward 15 which has the highest average household income, but is one of least populous wards, receives a lower budget allocation for Service Improvement and Enhancement projects, indicating that population density may be a more significant factor in determining funding levels than income levels.
 
@@
 
-### Relationship between Predictors and Budget Allocations (Growth-Related and State of Good Repair)
-
-[@fig-growth-related-budget]a) illustrates the relationship between average household income, population, and the total 10-year budget allocation for Growth-Related projects.
+### Relationship between Confounders and Budget Allocations (Growth-Related and State of Good Repair)
+
+[@fig-growth-related-budget]a) illustrates the relationship between average household income, population, and the total budget allocation for *Growth-Related projects*.
 The linear regression line for average household income remains relatively flat, hovering around \$100,000.
@@
 In contrast, the linear regression line for population shows a strong positive correlation with budget allocation for Growth-Related projects.
-Wards with higher population densities—such as Ward 10 (Spadina-Fort York), Ward 13 (Toronto Centre), Ward 5 (York South-Weston), and Ward 2 (Etobicoke North)—receive a larger share of the Growth-Related budget.
+Wards with higher population densities—such as Ward 10 (Spadina-Fort York) and Ward 3 (Toronto Centre)—receive a larger share of the Growth-Related budget.
 This indicates that population density plays an important role in determining funding levels for these projects, with more populous wards benefiting from increased allocations.
@@
 
-The purpose of this paper is to investigate the relationship between population density, average household income, and budget allocations across Toronto's 25 wards, and their potential impact on the economic well-being of each ward, as represented by the `total_building_permits` issued.
-To achieve this, we will first plot the data to visualize the relationships between these variables and the number of building permits issued in each ward independently.
-Then we will fit a Bayesian regression model with varying intercepts to estimate the effect of these variables on the number of building permits issued in each ward.
+The purpose of this paper is to investigate the relationship between population density, average household income, and budget allocations across Toronto's 25 wards, and their potential impact on the economic well-being of each ward, as represented by the `total_building_permits` issued, a indicator of infrastructure development.
+To achieve this, we will fit a Bayesian regression model with varying intercepts to estimate the effect of these variables on the number of building permits issued in each ward.
 
@@
 The standard deviation of the likelihood was set to an exponential distribution with a rate parameter of 1.
+
+\newpage
 
@@
 
-To better understand the underlying data, [@tbl-model-justification] provides a summary of the mean, median, and standard deviation for each variable used in the model.
-The predictor variables have large values and different scales, which can cause numerical instability and convergence issues in the model.
-To address this, we will standardize the predictor variables by subtracting the mean and dividing by the standard deviation.
-This will scale the variables consistently and improve both the model's convergence and interpretability.
-Additionally, we did not use a Bayesian Poisson regression model or negative binomial because the number of building permits is a continuous variable, not count data.
-Therefore, we chose to use a Bayesian multiple linear regression model.
-This model allows us to estimate the effect of population density, average household income, and budget allocations on the number of building permits issued in each ward giving us a well-rounded understanding of the relationships between these variables.
-
-The factors chosen for the model are known to influence the number of building permits issued in each ward.
-Population density and average household income are demographic and economic indicators that can affect construction activity.
-Budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects are key drivers of infrastructure development and resource distribution.
-Including these variables allows for assessing their individual and combined effects on the number of building permits issued.
-
-When population density was included, for the causual model, it exhibited exclusion restriction qualities, as it showed an unexpected negative relationship with building permits, leading to its exclusion for better interpretability.
-The model was then refitted without population density, focusing on average household income and budget allocations as predictors.
+To ensure model stability and interpretability, we standardized predictor variables (population density, average household income, and budget allocations) by subtracting their means and dividing by their standard deviations, as detailed in [@tbl-model-justification].
+This step addressed differences in scale and prevented numerical instability during Bayesian multiple linear regression modeling.
+
+The Bayesian multiple linear regression model was selected because the outcome variable—building permits—is continuous rather than count-based, making Poisson or negative binomial models unsuitable.
+The chosen predictors reflect demographic, economic, and infrastructural factors known to influence construction activity.
+Population density and average household income represent key socioeconomic indicators, while budget allocations for Growth-Related, State of Good Repair, and Service Improvement projects drive infrastructure investments.
+
+During causal analysis, population density exhibited unexpected exclusion restriction characteristics, showing a negative relationship with building permits.
+To enhance interpretability, population density was excluded from the model, leaving average household income and budget allocations as the primary predictors.
+The refitted model focuses on understanding how economic and infrastructural factors shape construction activity.
 
@@
 
-## Relationship between Predictor Variables and Building Permits
-
-We now explore the results of independent comparisons between the outcome variable (total number of building permits in Toronto's 25 wards) and the predictor variables (population density, average household income, and budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects).
-As shown in [@fig-building-permits-relationship], most predictor variables exhibit relatively flat linear regression lines, indicating a weak relationship with the total number of building permits.
-
-For example, the relationship between population density and building permits is notably flat, suggesting that population density has minimal influence on construction activity and development across the city’s wards—an unexpected result.
-A similar trend is observed for Growth-Related and State of Good Repair budget allocations, where flat regression lines suggest a weak relationship.
-This appears contradictory to earlier findings, where higher population density corresponded to increased budget allocations for these categories, which are typically associated with infrastructure development and maintenance.
-The same weak correlation is seen for Service Improvement and Enhancement budget allocations, as indicated by the flat regression line.
-
-In contrast, the relationship between average household income and building permits shows a strong positive correlation.
-Higher-income areas tend to experience more construction activity and development.
-For instance, Ward 15 (Don Valley West) has the highest number of building permits, aligning with its high average household income.
-This suggests that higher-income areas are more likely to see increased construction and development, highlighting a clear link between income levels and building permits.
+## Relationship between Confounders, Mediators and Budget Allocations {#sec-rel-budget}
+
+We now examine the results of independent comparisons between the outcome variable (total number of building permits in Toronto's 25 wards) and the predictor variables: population density, average household income, and budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects.
+As illustrated in [@fig-building-permits-relationship], most predictor variables demonstrate a strong positive correlation with the number of building permits, suggesting that higher population density, average household income, and budget allocations are generally associated with increased construction activity and urban development.
+
+However, the role of population density presents a different narrative.
+Although [@sec-rel-variables] revealed a positive correlation between population density and building permits, a counterintuitive negative relationship emerged within the model, leading to its exclusion.
+This anomaly is largely explained by the influence of Ward 10 (Spadina-Fort York) and Ward 3 (Etobicoke-Lakeshore), which consistently exhibited high population densities, substantial budget allocations, and elevated building permit counts.
+
+When these two wards are excluded, a clearer pattern emerges, Wards 8 (Eglinton-Lawrence), 11 (University-Rosedale), 2 (Etobicoke Centre), and 15 (Don Valley West) dominate in terms of building permits, average household income, and budget allocations.
+These findings suggest that, while population density plays a role in shaping construction activity, its effects are overshadowed by the more direct influence of income levels and budget distribution, warranting its exclusion to reduce model distortion and improve interpretability.
 
@@
 
-The results of the Bayesian regression model, summarized in [@tbl-model-results], provide important understanding into the relationship between total budget allocations, average household income, and the number of building permits issued across wards.
-
-The model explains 59.1% of the variance in building permits, indicating a moderate level of explanatory power.
-The coefficients suggest that both total budget (0.530) and average household income (0.350) have a positive and significant impact on the number of building permits issued, implying that areas with higher budget allocations and income levels are more likely to see increased construction activity.
-
-The inclusion of ward-specific random effects highlights the importance of accounting for variation between wards, suggesting that there are unobserved factors specific to each ward influencing permit issuance.
-Although the adjusted $R^2$ (0.369) indicates that the model's explanatory power is somewhat limited after accounting for the complexity of the predictors, the model still provides understanding into the role of economic factors in shaping construction activity.
-With an RMSE of 0.58, the model's predictions are reasonably accurate, although there is still room for improvement.
-
-In constructing the model, we assumed linearity and additive effects, acknowledging that these simplifications may not fully capture the interactions among variables.
-The model's limitations, such as the exclusion of population density due to its unexpected negative relationship, reflect a trade-off between interpretability and capturing complex dynamics.
-Nevertheless, the findings align with expectations that wealthier areas and those with greater budget allocations are more likely to experience construction growth, reaffirming the importance of economic resources in urban development patterns.
+The Bayesian regression model results, summarized in [@tbl-model-results], provide understanding into the relationship between budget allocations, average household income, and building permits across Toronto’s wards.
+The model explains 59.1% of the variance in building permits, with significant positive coefficients for total budget (0.525) and average household income (0.365).
+This indicates that wards with higher budgets and income levels are more likely to experience construction activity.
+
+Ward-specific random effects shows the influence of unobserved, ward-specific factors on permit issuance, while the adjusted ($R^2$) of 0.359 suggests moderate explanatory power after accounting for predictor complexity.
+With an RMSE of 0.57, the model achieves reasonable predictive accuracy.
+
+While the assumption of linearity and additive effects simplifies the analysis, it may overlook variable interactions.
+The exclusion of population density, due to its counterintuitive negative relationship, reflects a trade-off between clarity and capturing complex dynamics.
+Despite these limitations, the findings highlight the critical role of economic resources—both income and budget—in driving urban development, emphasizing the need for equitable resource allocation in shaping infrastructure growth.
 
@@
 As seen in [@sec-rel-variables] high-income areas often attract more construction activity, driven by market forces and residents’ ability to finance new developments.
-Whereas, only specific high-density areas like Ward 10 (Spadina-Fort York) and Ward 13 (Toronto Centre) receive substantial investment despite high population density, suggesting efforts to address this imbalance.
+Whereas, only specific high-density areas like Ward 10 (Spadina-Fort York) and Ward 13 (Toronto Centre) receive substantial investment despite high population density, justifying the reason for population density exclusion in the model.
 Meanwhile, as discussed in [@sec-distribution], low-density and lower-income wards risk being underserved.
@@
 
-## Survey, Sample, and Data Collection
-
-The relationship between income levels, population and construction outcomes is complex and influenced by diverse socioeconomic and regional factors.
-Current observational data, such as NHS survey to get income levels, and population of ward profiles but is inherently limited by its sample [@nhs_canada].
-To better answer our research question, a more targeted survey approach is essential.
-This exploration outlines an idealized survey methodology designed to reduce bias, ensure representativeness, and deepen understanding.
-
-### Idealized Survey Design for Construction Analysis 
-A survey specifically designed to explore the link between income and construction outcomes would involve the following key components:
-
-a. Defining the Survey Objectives
-The survey would aim to investigate how income disparities influence construction investment, access to resources, regulatory hurdles, and housing quality across different regions.
-
-b. Sampling Methodology  
-To ensure a representative and unbiased dataset, the survey would employ:
-
-- Stratified Sampling: Income levels and ward classifications (e.g., urban, suburban, rural) would be stratified to ensure proportional representation.  
-- Random Sampling Within Strata: Participants would be randomly selected within each stratum to enhance the validity and generalizability of the findings.
-
-c. Sample Size and Distribution
-The sample size should be sufficiently large to allow for statistically significant comparisons across different income levels and geographic regions.
-A minimum target of several thousand participants distributed evenly across strata would be ideal.
-
-### Reducing Bias in Sampling and Data Collection  
-Bias reduction is important in survey design to ensure reliable and actionable understanding. This involves addressing various types of bias:
-
-a. Sampling Bias Mitigation
-Stratified random sampling ensures that underrepresented groups, particularly those in lower income brackets or remote regions, are adequately included.
-
-b. Non-Response Bias Mitigation 
-Offering multiple survey delivery methods—such as online platforms, in-person interviews, and mailed surveys—enhances accessibility and participation.
-Additionally, providing incentives like small monetary rewards or gift cards can boost response rates among diverse groups.
-
-c. Design Bias Mitigation 
-To avoid influencing responses, survey questions must be neutrally worded and pilot-tested to ensure clarity.
-Cognitive interviews with a subset of participants can help refine the wording and structure of questions.
+The relationship between income levels, population density, and construction outcomes reflects complex interactions shaped by diverse socioeconomic and regional factors.
+Existing datasets, such as the NHS survey, provide a starting point but do not fully capture these dynamics.
+To address this, a well-structured survey methodology is outlined to ensure greater representativeness and reliability.
+
+## Survey Design for Construction Analysis
+
+The survey seeks to understand how income levels influence construction investments, access to resources, regulatory challenges, and housing quality across geographic regions.
+This allows for a detailed assessment of regional and economic differences in construction outcomes.
+
+## Sampling Methodology
+The proposed design prioritizes representativeness through the following methods:
+
+- Stratified Sampling: Groups are categorized by income levels and ward classifications (high-income, populated, and etc.) to ensure diverse representation.  
+- Random Sampling Within Strata: This approach minimizes selection bias, ensuring participants across income groups are adequately represented.
+
+## Ensuring Robust Data Collection
+
+- Sampling Bias: Using stratified random sampling ensures inclusion of marginalized groups, particularly those in remote wards or lower income brackets, often underrepresented in traditional data sources.
+- Non-Response Bias: Offering multiple delivery modes—such as online forms, mailed surveys, and in-person interviews—accommodates varying levels of accessibility.
+To encourage participation, incentives like gift cards or small monetary rewards can be provided.
+- Question Design: Survey questions must be neutrally worded to avoid influencing responses. Pilot testing and iterative feedback from participants will refine question clarity and relevance.  
+- Human-Centric Approach: Present the survey as an opportunity for participants to share their experiences and shape future construction planning.  
+- Visual Enhancements: Include infographics or interactive components to make participation more engaging and relatable.  
+- Validation Through Multiple Sources: To strengthen findings, the survey data should be cross-referenced with secondary sources, such as municipal reports, and supplemented with qualitative interviews where necessary.
+
+This approach ensures a structured methodology that balances statistical rigor with participant-focused design, allowing for a meaningful exploration of how socioeconomic factors influence construction outcomes across regions.
 

### 2024-12-15 paper/paper.qmd
Grammer Fix
- Fixed grammer

@@
 title: "Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto"
-subtitle: "Population density drive larger budget allocations meanwhile average household income drives construction projects"
+subtitle: "Population density influences larger budget allocations, while average household income determines the focus of construction projects"
 author: 
@@
 [@sec-discussion] explores the implications for urban governance and budget planning in Toronto.
-Finally, [@sec-limitations] addresses the study’s limitations and offers recommendations for future research.
+Finally, [@sec-limitations] addresses the study’s limitations and suggests recommendations for future research.
 Additional details, including tables, figures, and methodology, are provided in the [@sec-appendix].
@@
 Toronto Center (Ward 13), Don Valley East (Ward 16), and Etobicoke Center (Ward 2) also received notable budget allocations, some of these wards have higher average household incomes as well as higher population densities.
-The box plots reveal that budget allocations for *Growth Related* projects have increased steadily over the years, with 75% of wards receiving higher allocations in 2024 compared to 2021.
+The box plots presents that budget allocations for *Growth Related* projects have increased steadily over the years, with 75% of wards receiving higher allocations in 2024 compared to 2021.
 In contrast, budget allocations for *State of Good Repair* and *Service Improvement and Enhancement* projects have remained relatively consistent, with few outliers receiving higher budget allocations.
@@
 This analysis has several limitations that warrant consideration.
-As discussed in [@sec-data], the datasets used, sourced from the City of Toronto, provide insights into the city’s demographics, budget allocations, and building permits.
+As discussed in [@sec-data], the datasets used, sourced from the City of Toronto, presents city’s demographics, budget allocations, and building permits.
 However, they may not fully capture the complexity of Toronto's infrastructure development and resource distribution.
@@
 This narrow focus may overlook important stages in the construction process.
-Expanding the analysis to include more permit statuses could offer a more comprehensive understanding of construction activity.
-
-Another limitation is the lack of detailed data on specific projects within each budget category, which restricts insights into why certain wards receive more or less funding.
+Expanding the analysis to include more permit statuses could offer a more in-depth understanding of construction activity.
+
+Another limitation is the lack of detailed data on specific projects within each budget category, which restricts the paper's analysis into why certain wards receive more or less funding.
 Access to granular project-level data could provide deeper context and explanations for funding disparities.
